---
title: "DuckDB for Marketing Analytics: From Parquet to Powerful Queries"
subtitle: "How to Combine DuckDB with Arrow + Parquet"
description: "This page describes DuckDB and compares it with Arrow and Parquet and explain how to use them together effectively when dealing with big data"
categories: [DuckDB, Arrow, Parquet]
order: 2
author: "Jae Jung"
date: today
format: 
    html:
        theme: cosmo
        toc: true
        toc-depth: 3
        toc-expand: 1
        code-fold: show
        toc-location: right-body
        number-sections: true
        #embed-resources: true
execute:
  warning: false
  message: false
  echo: fenced
  cache: true
  freeze: auto
---

# Introduction

## What is DuckDB?

DuckDB is an embedded analytical database engine — think of it as "SQLite, but built for analytics." It runs entirely inside your R process with no server, no installation, no configuration. You just install the package and it works.

It was designed from the ground up for **OLAP** (Online Analytical Processing) workloads — the kind of heavy aggregations, joins, and scans that are typical in data analysis, as opposed to transactional row-by-row operations.

-   SQLite was built for this (transactional — many small reads/writes):
    -   "Insert one customer order"
    -   "Update one user's address"
    -   "Delete one session record"
-   DuckDB was built for this (analytical — few large scans):
    -   "What was total revenue by channel across 50 million events?"
    -   "Which pages had the highest bounce rate last quarter?"
    -   "Give me the 99th percentile session duration by device type"

## Three Remarkable Features

### 1. It Lives Inside R — No Server Needed

**Traditional database setup** (PostgreSQL, MySQL, Snowflake):

1.  Install database server
2.  Configure network/ports
3.  Create users and permissions
4.  Start the server process
5.  Connect from R
6.  Manage the server forever

**DuckDB setup**:

```         
install.packages("duckdb")
library(duckdb)
con <- dbConnect(duckdb::duckdb())   # That's it. Running in 1 second.
```

### 2. It Is Shockingly Fast

> DuckDB uses columnar storage and vectorised execution — it reads only the columns your query needs and processes thousands of rows simultaneously using modern CPU instructions:

**Benchmark:** Count events by type across 10 million GA4 rows

-   data.frame in RAM: \~4.2 seconds
-   data.table: \~0.8 seconds
-   dplyr on tibble: \~3.1 seconds
-   DuckDB on parquet file: \~0.3 seconds ← faster than data in RAM
-   DuckDB on DuckDB file: \~0.1 seconds

### 3. It Reads Parquet Directly — Without Loading Into RAM

> This is the feature that changes everything for large datasets:

```{{r}}
# Traditional approach — must fit in RAM
ga4_data <- arrow::read_parquet("data/ga4_events.parquet")  # ~4 GB into RAM
result   <- ga4_data |> filter(...) |> summarise(...)        # then query

# DuckDB approach — reads only what the query needs
con    <- dbConnect(duckdb::duckdb())
result <- dbGetQuery(con, "
  SELECT event_name, COUNT(*) as n
  FROM read_parquet('data/ga4_events.parquet')
  WHERE event_date >= '2021-01-01'
  GROUP BY event_name
")
# DuckDB scans only the event_name and event_date columns
# Never loads price, device, geo, or any other column into RAM
```

## Why use DuckDB?

**It's fast.** DuckDB uses a columnar execution engine with vectorized query processing, meaning it processes chunks of columns at a time rather than row by row. For analytical queries on large data, this is dramatically faster than base R or even many database solutions.

**It's serverless.** No Docker, no Postgres instance, no DBA needed. It lives inside your session like any other R package.

**It speaks SQL fluently.** You can write standard SQL, or use it as a dplyr backend via dbplyr — so you don't need to learn a new API.

**It handles larger-than-memory data.** DuckDB can query files that don't fit in RAM by streaming and processing them in chunks, making it genuinely useful for big data workflows.

**It integrates natively with Parquet and Arrow.** It can directly query `.parquet` files on disk without loading them into memory first — no import step needed.

## DuckDB vs Arrow vs Parquet

These three are complementary, not competing. Understanding the role of each is important:

|   | **Parquet** | **Arrow** | **DuckDB** |
|------------------|------------------|------------------|------------------|
| What is it? | File format (How data is stored on disk) | In-memory format + compute engine (a toolkit): how data is held in RAM and processed | A query engine (how you ask questions of data) |
| Lives where? | Disk | RAM | RAM + disk  (processes from disk) |
| Primary role | Store data efficiently | Process data in memory | Query data with SQL/dplyr" |
| Reads Parquet | IS Parquet | Yes — natively | Yes — directly |
| SQL support | No | Limited (via DuckDB) | Full SQL |
| dplyr support | Via arrow package | Yes — first class" | Yes — via dbplyr |
| Larger-than-memory? | Yes (it's on disk) | Partial (via chunked datasets) | Yes – streams from disk |
| Query language | None | dplyr verbs | SQL + dplyr via dbplyr |
| Speed (analytics) | N/A (Storage only) | Very fast | Extremely fast |
| Best for | Archiving & sharing data | In-memory transformation: e.g., Wrangling in R/Python | Complex queries, joins, aggregations |

The typical mental model is:

-   **Parquet** is where your data *lives* (on disk, compressed, efficient)

-   **Arrow** is how you *move and process* it quickly in memory

-   **DuckDB** is how you *query* it with full SQL power, with or without loading into memory

They work beautifully together — DuckDB can read Arrow tables directly and query Parquet files on disk, and you can pass results back to Arrow or collect into R.

# How DuckDB Works with Arrow and Parquet

There are **four integration patterns**, each suited to different situations:

## Setup

```{{r}}
library(duckdb)
library(arrow)
library(dplyr)
```

## Pattern 1. Query a Parquet file directly with DuckDB (no loading into R)

-   The most direct approach — point DuckDB at a Parquet file and query it with SQL, no loading required:

-   DuckDB can scan Parquet files on disk using SQL — nothing enters R memory until you `collect()`.

-  **Simple**:

```{r}
#| eval: false

con <- dbConnect(duckdb())

# Register and query a parquet file directly
dbGetQuery(con, "
  SELECT region, SUM(sales) AS total_sales
  FROM read_parquet('data/sales.parquet')
  GROUP BY region
  ORDER BY total_sales DESC
")
```

-  **Advanced**:

```{r}
#| eval: false

library(duckdb)
library(DBI)
con <- dbConnect(duckdb())

# Register and query a parquet file directly
dbGetQuery(
  con,
  "
  SELECT region, SUM(sales) AS total_sales
  FROM read_parquet('data/sales.parquet')
  GROUP BY region
  ORDER BY total_sales DESC
"
)

# ── Query a single Parquet file ────────────────────────────────────────────────
dbGetQuery(
  con,
  "
  SELECT
    event_name,
    COUNT(*) AS events,
    COUNT(DISTINCT user_pseudo_id) AS users
  FROM read_parquet('data/ga4_flat.parquet')
  GROUP BY event_name
  ORDER BY events DESC
  LIMIT 10
"
)

# ── Query multiple Parquet files with a wildcard ───────────────────────────────
# Perfect for GA4's one-file-per-day structure
dbGetQuery(
  con,
  "
  SELECT
    event_date,
    COUNT(*) AS daily_events
  FROM read_parquet('data/ga4_*.parquet')
  GROUP BY event_date
  ORDER BY event_date
"
)

dbGetQuery(
  con,
  "
  SELECT year, SUM(revenue) AS total
  FROM read_parquet('data/sales/*.parquet', union_by_name = TRUE)
  GROUP BY year
  ORDER BY year
"

# ── Join two Parquet files in a single query ───────────────────────────────────
dbGetQuery(
  con,
  "
  SELECT
    s.device_category,
    s.traffic_source_medium,
    COUNT(DISTINCT e.user_pseudo_id) AS users,
    SUM(CASE WHEN e.event_name = 'purchase' THEN 1 ELSE 0 END) AS purchases
  FROM read_parquet('data/ga4_flat.parquet')    AS e
  JOIN read_parquet('data/ga4_sessions.parquet') AS s
    ON e.user_pseudo_id = s.user_pseudo_id
   AND e.event_date     = s.event_date
  GROUP BY s.device_category, s.traffic_source_medium
  ORDER BY users DESC
"
)

dbDisconnect(con, shutdown = TRUE)
```

This is powerful — DuckDB pushes the entire query down to the file, reading only the columns and rows it needs (thanks to Parquet's columnar layout).

## Parttern 2. Use DuckDB as a dplyr backend (no SQL required)

> If you prefer dplyr verbs over SQL, you can register the **Parquet file** as a **virtual table** and query it lazily.

> `dbplyr` translates dplyr verbs into DuckDB SQL automatically:

-  **Simple**:

```{r}
#| eval: false
con <- dbConnect(duckdb())

duckdb_register(con, "sales", read_parquet("data/sales.parquet"))

tbl(con, "sales") |>
  filter(year >= 2022) |>
  group_by(region) |>
  summarise(total = sum(sales)) |>
  arrange(desc(total)) |>
  collect()   # only now does data enter R
```

-  **Advanced**:
```{r}
#| eval: false
library(duckdb)
library(dplyr)
library(dbplyr)

con <- dbConnect(duckdb())

# ── Register Parquet file as a virtual table ───────────────────────────────────
# This creates a view — no data is copied or loaded
duckdb_read_csv   # (there is also this for CSV)
dbExecute(con, "
  CREATE VIEW ga4_events AS
  SELECT * FROM read_parquet('data/ga4_flat.parquet')
")

dbExecute(con, "
  CREATE VIEW ga4_sessions AS
  SELECT * FROM read_parquet('data/ga4_sessions.parquet')
")

# ── Now use dplyr as if it were a regular data frame ──────────────────────────
events   <- tbl(con, "ga4_events")
sessions <- tbl(con, "ga4_sessions")

# dplyr query — translated to DuckDB SQL by dbplyr
top_events <- events |>
  count(event_name, sort = TRUE) |>
  head(10) |>
  collect()    # collect() executes the query and returns a tibble

print(top_events)

# Join two Parquet files using dplyr — no SQL written
device_conversions <- sessions |>
  select(user_pseudo_id, device_category, traffic_source_medium) |>
  left_join(
    events |>
      filter(event_name == "purchase") |>
      distinct(user_pseudo_id) |>
      mutate(purchased = TRUE),
    by = "user_pseudo_id"
  ) |>
  mutate(purchased = coalesce(purchased, FALSE)) |>
  group_by(device_category, traffic_source_medium) |>
  summarise(
    users      = n_distinct(user_pseudo_id),
    purchasers = sum(as.integer(purchased), na.rm = TRUE),
    cvr        = purchasers / users,
    .groups    = "drop"
  ) |>
  arrange(desc(cvr)) |>
  collect()  # only now does data enter R

print(device_conversions)

dbDisconnect(con, shutdown = TRUE)

```

## Pattern 3. Query an Arrow Table with DuckDB (The Power Combination)

> Arrow and DuckDB are designed to work together. Arrow handles file I/O and in-memory transformation; DuckDB handles querying. 

> They share data without copying using the Apache Arrow memory format.

> If you already have an Arrow table in memory, DuckDB can query it directly — zero copying involved:

-  **Simple**:
```{r}
#| eval: false

# Create or load an Arrow table
arrow_table <- read_parquet("data/sales.parquet", as_data_frame = FALSE)

con <- dbConnect(duckdb())
duckdb_register_arrow(con, "sales_arrow", arrow_table)

tbl(con, "sales_arrow") |>
  filter(category == "Electronics") |>
  summarise(avg_price = mean(price)) |>
  collect()
```

- **Advanced**:

```{r}
#| eval: false

library(arrow)
library(duckdb)
library(dplyr)

# ── Step 1: Open Parquet as Arrow Dataset (lazy — nothing loaded yet) ──────────
ga4_arrow <- arrow::open_dataset("data/ga4_flat.parquet")

# ── Step 2: Do transformations in Arrow ───────────────────────────────────────
# Arrow is better at: filtering rows, selecting columns, type conversion
ga4_prepared <- ga4_arrow |>
  filter(event_name %in% c("purchase", "add_to_cart", "view_item")) |>
  select(event_date, event_name, user_pseudo_id) |>
  mutate(event_date = as.Date(event_date))

# ── Step 3: Hand off to DuckDB for aggregation ────────────────────────────────
# DuckDB is better at: GROUP BY, window functions, complex joins
con <- dbConnect(duckdb::duckdb())

# Register the Arrow object directly — ZERO data copying
duckdb::duckdb_register_arrow(con, "ga4_prepared", ga4_prepared)

# Now query it with full DuckDB SQL power
result <- dbGetQuery(con, "
  SELECT
    event_date,
    event_name,
    COUNT(DISTINCT user_pseudo_id) AS users
  FROM ga4_prepared
  GROUP BY event_date, event_name
  ORDER BY event_date, event_name
")

print(result)
dbDisconnect(con, shutdown = TRUE)

# ── The same workflow using dplyr throughout ───────────────────────────────────
library(arrow)
library(duckdb)
library(dplyr)

con <- dbConnect(duckdb::duckdb())

# Arrow dataset → register with DuckDB → query with dplyr
ga4_arrow <- arrow::open_dataset("data/ga4_flat.parquet")
duckdb::duckdb_register_arrow(con, "ga4_events", ga4_arrow)

ga4_params_arrow <- arrow::open_dataset("data/ga4_event_params.parquet")
duckdb::duckdb_register_arrow(con, "ga4_params", ga4_params_arrow)

# Now use pure dplyr — no SQL needed at all
# dbplyr translates these verbs to DuckDB SQL automatically

# Marketing analysis: page views by device and channel
page_analysis <- tbl(con, "ga4_params") |>
  filter(key == "page_location") |>
  rename(page_location = param_value) |>
  left_join(
    tbl(con, "ga4_events") |>
      select(event_timestamp, user_pseudo_id,
             device_category, traffic_source_medium),
    by = c("event_timestamp", "user_pseudo_id")
  ) |>
  mutate(
    page_path = regexp_replace(page_location, "\\?.*$", ""),
    page_path = regexp_replace(page_path, "https?://[^/]+", "")
  ) |>
  group_by(device_category, page_path) |>
  summarise(
    page_views = n(),
    users      = n_distinct(user_pseudo_id),
    .groups    = "drop"
  ) |>
  filter(page_views >= 10) |>
  arrange(desc(page_views)) |>
  head(20) |>
  collect()

print(page_analysis)
dbDisconnect(con, shutdown = TRUE)
```

## Pattern 4. Persistent DuckDB Database (Best for a Course Project)

> Write your rectangled GA4 tables once into a `.duckdb` file. Every subsequent analysis opens this one file — no BigQuery, no Parquet management, no RAM issues:

```{r}
#| eval: false

library(duckdb)
library(dplyr)
library(arrow)

# ── SETUP: Run once to create the database ────────────────────────────────────
con <- dbConnect(
  duckdb::duckdb(),
  dbdir     = "data/ga4.duckdb",
  read_only = FALSE
)

# Write tables from Parquet files — DuckDB reads and stores them
dbExecute(con, "
  CREATE TABLE ga4_events AS
  SELECT * FROM read_parquet('data/ga4_flat.parquet')
")

dbExecute(con, "
  CREATE TABLE ga4_params AS
  SELECT * FROM read_parquet('data/ga4_event_params.parquet')
")

dbExecute(con, "
  CREATE TABLE ga4_sessions AS
  SELECT * FROM read_parquet('data/ga4_sessions.parquet')
")

dbExecute(con, "
  CREATE TABLE ga4_items AS
  SELECT * FROM read_parquet('data/ga4_items.parquet')
")

# Add indexes for faster querying
dbExecute(con, "CREATE INDEX idx_event_name ON ga4_events(event_name)")
dbExecute(con, "CREATE INDEX idx_param_key  ON ga4_params(key)")
dbExecute(con, "CREATE INDEX idx_event_date ON ga4_events(event_date)")

# Confirm what was saved
cat("Tables in ga4.duckdb:\n")
dbListTables(con) |> print()

cat("\nRow counts:\n")
purrr::map_df(dbListTables(con), \(t) {
  tibble::tibble(
    table = t,
    rows  = dbGetQuery(con,
              glue::glue("SELECT COUNT(*) n FROM {t}"))$n
  )
}) |> print()

cat("\nFile size:",
    round(file.size("data/ga4.duckdb") / 1024^2, 1), "MB\n")

dbDisconnect(con, shutdown = TRUE)
```

```{r}
#| eval: false

# ── ANALYSIS: Open the database for every subsequent session ──────────────────
con <- dbConnect(
  duckdb::duckdb(),
  dbdir     = "data/ga4.duckdb",
  read_only = TRUE    # Prevents accidental writes during analysis
)

# Lazy table references — nothing in RAM yet
events   <- tbl(con, "ga4_events")
params   <- tbl(con, "ga4_params")
sessions <- tbl(con, "ga4_sessions")
items    <- tbl(con, "ga4_items")

# ── Example 1: Event counts (dplyr) ───────────────────────────────────────────
events |>
  count(event_name, sort = TRUE) |>
  collect()

# ── Example 2: Top pages (dplyr) ──────────────────────────────────────────────
params |>
  filter(key == "page_location") |>
  count(param_value, sort = TRUE) |>
  head(10) |>
  collect()

# ── Example 3: Revenue by day (SQL) ───────────────────────────────────────────
dbGetQuery(con, "
  SELECT
    event_date,
    COUNT(*)                                   AS transactions,
    SUM(ecommerce_purchase_revenue_in_usd)     AS revenue,
    AVG(ecommerce_purchase_revenue_in_usd)     AS avg_order_value
  FROM ga4_events
  WHERE event_name = 'purchase'
  GROUP BY event_date
  ORDER BY event_date
")

# ── Example 4: Funnel analysis (SQL window functions) ─────────────────────────
dbGetQuery(con, "
  WITH funnel AS (
    SELECT
      event_name,
      COUNT(DISTINCT user_pseudo_id) AS users,
      ROW_NUMBER() OVER (
        ORDER BY CASE event_name
          WHEN 'view_item'       THEN 1
          WHEN 'add_to_cart'     THEN 2
          WHEN 'begin_checkout'  THEN 3
          WHEN 'purchase'        THEN 4
        END
      ) AS stage_order
    FROM ga4_events
    WHERE event_name IN ('view_item','add_to_cart',
                         'begin_checkout','purchase')
    GROUP BY event_name
  )
  SELECT
    event_name,
    users,
    ROUND(100.0 * users /
      FIRST_VALUE(users) OVER (ORDER BY stage_order), 1) AS pct_of_top
  FROM funnel
  ORDER BY stage_order
")

dbDisconnect(con, shutdown = TRUE)
```

# How the Three Tools Form a Complete Stack

```
# The complete GA4 analytics stack:

# ┌─────────────────────────────────────────────────────────────┐
# │                     YOUR R SESSION                          │
# │                                                             │
# │  ┌──────────┐    ┌──────────┐    ┌──────────────────────┐  │
# │  │  Arrow   │    │  dplyr   │    │        ggplot2        │  │
# │  │          │    │  dbplyr  │    │        gt             │  │
# │  │ Read &   │    │          │    │                       │  │
# │  │ Write    │    │ Verbs →  │    │  Visualise & Report   │  │
# │  │ Parquet  │    │ SQL      │    │                       │  │
# │  └────┬─────┘    └────┬─────┘    └───────────┬──────────┘  │
# │       │               │                       │             │
# │       └───────────────▼───────────────────────┘             │
# │                  ┌──────────┐                               │
# │                  │  DuckDB  │                               │
# │                  │          │                               │
# │                  │  Query   │                               │
# │                  │  Engine  │                               │
# │                  └────┬─────┘                               │
# └───────────────────────┼─────────────────────────────────────┘
#                         │
#          ┌──────────────┼──────────────┐
#          │              │              │
#    ┌─────▼──────┐ ┌─────▼──────┐ ┌───▼────────────┐
#    │  Parquet   │ │  Parquet   │ │  ga4.duckdb    │
#    │  ga4_flat  │ │  ga4_params│ │  (all tables)  │
#    └────────────┘ └────────────┘ └────────────────┘
#         DISK            DISK           DISK

# Arrow   = the loading dock (moves data efficiently)
# DuckDB  = the analyst (asks and answers questions)
# Parquet = the filing system (stores data efficiently)

```

# The Big Picture Workflow

```{r}
#| echo: false
#install.packages("DiagrammeR")
library(DiagrammeR)

grViz("
  digraph workflow {
    graph [layout = dot, rankdir = TD]
    node [shape = rectangle, style = filled]
    
    A [label = 'Parquet files on disk', fillcolor = '#C0DD97']
    B [label = 'DuckDB queries', fillcolor = '#B5D4F4']
    C [label = 'Large intermediate result?', fillcolor = '#FAC775', shape = diamond]
    D [label = 'compute()', fillcolor = '#B5D4F4']
    E [label = 'collect()', fillcolor = '#B5D4F4']
    F [label = 'Visualize - ggplot2', fillcolor = '#F4C0D1']
    G [label = 'Model - tidymodels', fillcolor = '#F4C0D1']
    
    A -> B -> C
    C -> D [label = 'Yes']
    C -> E [label = 'No']
    D -> E -> F
    E -> G
  }
")
```

**Parquet is your warehouse, Arrow is your forklift, DuckDB is your analyst.** Each does its job, and they hand off to each other seamlessly.