---
title: "DuckDB for Marketing Analytics: From Parquet to Powerful Queries"
subtitle: "How to Combine DuckDB with Arrow + Parquet"
description: "This page describes DuckDB and compares it with Arrow and Parquet and explain how to use them together effectively when dealing with big data"
categories: [DuckDB, Arrow, Parquet]
order: 2
author: "Jae Jung"
date: today
format: 
    html:
        theme: cosmo
        toc: true
        toc-depth: 3
        toc-expand: 1
        code-fold: show
        toc-location: right-body
        number-sections: true
        #embed-resources: true
execute:
  warning: false
  message: false
  echo: fenced
  cache: true
  freeze: auto
---

# Introduction

## What is DuckDB?

DuckDB is an embedded analytical database engine — think of it as "SQLite, but built for analytics." It runs entirely inside your R process with no server, no installation, no configuration. You just install the package and it works.

It was designed from the ground up for **OLAP** (Online Analytical Processing) workloads — the kind of heavy aggregations, joins, and scans that are typical in data analysis, as opposed to transactional row-by-row operations.

-   SQLite was built for this (transactional — many small reads/writes):
    -   "Insert one customer order"
    -   "Update one user's address"
    -   "Delete one session record"
-   DuckDB was built for this (analytical — few large scans):
    -   "What was total revenue by channel across 50 million events?"
    -   "Which pages had the highest bounce rate last quarter?"
    -   "Give me the 99th percentile session duration by device type"

## Three Remarkable Features

### 1. It Lives Inside R — No Server Needed

**Traditional database setup** (PostgreSQL, MySQL, Snowflake):

1.  Install database server
2.  Configure network/ports
3.  Create users and permissions
4.  Start the server process
5.  Connect from R
6.  Manage the server forever

**DuckDB setup**:

```{r}
#| eval: false         
install.packages("duckdb")
library(duckdb)
con <- dbConnect(duckdb())   # That's it. Running in 1 second.
```

### 2. It Is Shockingly Fast

> DuckDB uses columnar storage and vectorised execution — it reads only the columns your query needs and processes thousands of rows simultaneously using modern CPU instructions:

**Benchmark:** Count events by type across 10 million GA4 rows

-   data.frame in RAM: \~4.2 seconds
-   data.table: \~0.8 seconds
-   dplyr on tibble: \~3.1 seconds
-   DuckDB on parquet file: \~0.3 seconds ← faster than data in RAM
-   DuckDB on DuckDB file: \~0.1 seconds

### 3. It Reads Parquet Directly — Without Loading Into RAM

> This is the feature that changes everything for large datasets:

```{{r}}
# Traditional approach — must fit in RAM
ga4_data <- arrow::read_parquet("data/ga4_events.parquet")  # ~4 GB into RAM
result   <- ga4_data |> filter(...) |> summarise(...)        # then query

# DuckDB approach — reads only what the query needs
con    <- dbConnect(duckdb::duckdb())
result <- dbGetQuery(con, "
  SELECT event_name, COUNT(*) as n
  FROM read_parquet('data/ga4_events.parquet')
  WHERE event_date >= '2021-01-01'
  GROUP BY event_name
")
# DuckDB scans only the event_name and event_date columns
# Never loads price, device, geo, or any other column into RAM
```

## Why use DuckDB?

1.  **It's fast.** DuckDB uses a columnar execution engine with vectorized query processing, meaning it processes chunks of columns at a time rather than row by row. For analytical queries on large data, this is dramatically faster than base R or even many database solutions.

2.  **It's serverless.** No Docker, no Postgres instance, no DBA needed. It lives inside your session like any other R package.

3.  **It speaks SQL fluently.** You can write standard SQL, or use it as a dplyr backend via dbplyr — so you don't need to learn a new API.

4.  **It handles larger-than-memory data.** DuckDB can query files that don't fit in RAM by streaming and processing them in chunks, making it genuinely useful for big data workflows.

5.  **It integrates natively with Parquet and Arrow.** It can directly query `.parquet` files on disk without loading them into memory first — no import step needed.

## DuckDB vs Arrow vs Parquet

> These three are complementary, not competing. Understanding the role of each is important:

|   | **Parquet** | **Arrow** | **DuckDB** |
|------------------|------------------|------------------|------------------|
| What is it? | File format (How data is stored on disk) | In-memory format + compute engine (a toolkit): how data is held in RAM and processed | A query engine (how you ask questions of data) |
| Lives where? | Disk | RAM | RAM + disk (processes from disk) |
| Primary role | Store data efficiently | Process data in memory | Query data with SQL/dplyr |
| Reads Parquet | IS Parquet | Yes — natively | Yes — directly |
| SQL support | No | Limited (via DuckDB) | Full SQL |
| dplyr support | Via arrow package | Yes — first class" | Yes — via dbplyr |
| Larger-than-memory? | Yes (it's on disk) | Partial (via chunked datasets) | Yes – streams from disk |
| Query language | None | dplyr verbs | SQL + dplyr via dbplyr |
| Speed (analytics) | N/A (Storage only) | Very fast | Extremely fast |
| Best for | Archiving & sharing data | In-memory transformation: e.g., Wrangling in R/Python | Complex queries, joins, aggregations |

> The typical mental model is:

-   **Parquet** is where your data *lives* (on disk, compressed, efficient)

-   **Arrow** is how you *move and process* it quickly in memory

-   **DuckDB** is how you *query* it with full SQL power, with or without loading into memory

::: callout-tip
They work beautifully together — DuckDB can read Arrow tables directly and query Parquet files on disk, and you can pass results back to Arrow or collect into R.
:::

# Setup: Connecting to BigQuery

## Authentication and Connection

::: callout-important
## Before You Begin

You need a **Google Cloud Platform (GCP) account** with a project that has billing enabled. Querying the public dataset is free in terms of data costs (you won't be charged for the data itself), but GCP bills for query processing. Small queries on this dataset cost fractions of a cent.

1.  Go to [console.cloud.google.com](https://console.cloud.google.com)
2.  Create a project (or use an existing one)
3.  Enable the BigQuery API for that project
4.  Copy your **Project ID** — you'll need it below
:::

```{r}
#| label: connection
#| eval: false

# ── Replace with your GCP Project ID ─────────────────────────────────────────
# Either set the environment variable: usethis::edit_r_environ()
# and add: GCP_PROJECT_ID=your-actual-project-id
# Or assign directly (less portable):
billing_project <- "GCP_PROJECT_ID"

# ── Authenticate ──────────────────────────────────────────────────────────────
# First run: opens a browser for OAuth. Subsequent runs: uses cached token.
# In Positron IDE, if the browser redirect fails, add: use_oob = TRUE
bq_auth()
# bq_auth(use_oob = TRUE)  # Uncomment if you see an OAuth redirect error

# ── Connect to the public GA4 dataset ─────────────────────────────────────────
con <- dbConnect(
  bigrquery::bigquery(),
  project = "bigquery-public-data",
  dataset = "ga4_obfuscated_sample_ecommerce",
  billing = billing_project
)

# Verify connection: list available tables
dbListTables(con)
```

## Downloading and Caching the Data

> GA4 tables are large. For this workshop we download **one week of events** (\~150,000 rows) once, then save locally to avoid re-querying BigQuery on every render. 

```{r}
#| label: download-data
#| eval: false

# ── Run this chunk ONCE. It saves data locally. ───────────────────────────────
dir.create("data", showWarnings = FALSE)

query_result <- bq_project_query(
  billing_project,
  "SELECT
     event_date,
     event_timestamp,
     event_name,
     event_params,
     user_pseudo_id,
     device,
     geo,
     traffic_source,
     ecommerce,
     items
   FROM `bigquery-public-data.ga4_obfuscated_sample_ecommerce.events_*`
   WHERE _TABLE_SUFFIX BETWEEN '20210101' AND '20210107'" # 90 days of data
)

# A tibble data
ga4_raw <- query_result |>
  bq_table_download(
    bigint = "numeric", # Convert INT64 to R numeric (sufficient for most analyses)
    page_size = 1000 # Download in batches of 10k rows
  )
```

:::callout-tip
Since the data is relatively small size, we have three options to save the data as you can see below

-   RDS, 
-   One Parquet file, 
-   Partitioned Parquet files.
:::

### Option 1: Saving it as RDS file

```{r}
#| label: save-rds
#| eval: false
saveRDS(ga4_raw, "data/ga4_raw.rds")
message("Saved ", nrow(ga4_raw), " rows to data/ga4_raw.rds")
```

> **Note**: RDS stands for R Data Serialization

```{r}
#| label: load-rds
#| eval: false

# ── Load from local cache (use this for all subsequent analyses) ──────────────
ga4 <- readRDS("data/ga4_raw.rds")
```

### Option 2: Saving as a single Parquet file

```{r}
#| label: save-parquet
#| eval: false

library(arrow)

write_parquet(ga4_raw, "data/ga4_raw.parquet")
message("Saved to data/ga4_raw.parquet")
```
```{r}
#| label: load-parquet
#| eval: false

ga4 <- read_parquet("data/ga4_raw.parquet")
```

### Options 3: Saving as Partitioned Parquet

> Also no iteration needed — write_dataset() handles the partitioning automatically:
```{r}
#| label: save-partitioned
#| eval: false

library(arrow)
library(dplyr)

ga4_raw |>
  group_by(event_date) |>
  write_dataset("data/ga4_parquet", format = "parquet")

message("Saved partitioned Parquet to data/ga4_parquet/")
```

> **Note**: `write_dataset()` automatically creates one subfolder per `event_date`,
> e.g. `data/ga4_parquet/event_date=20210101/part-0.parquet`

```{r}
#| label: load-partitioned
#| eval: false

ga4 <- open_dataset("data/ga4_parquet/")
```

# How DuckDB Works with Arrow and Parquet

> There are **four integration patterns**, each suited to different situations:

**Setup:**

```{{r}}
library(duckdb)
library(arrow)
library(dplyr)
```

## Pattern 1. Query a Parquet file directly with DuckDB (no loading into R)

-   The most direct approach — point DuckDB at a `Parquet` file and query it with SQL, no loading required:

-   DuckDB can scan Parquet files on disk using `SQL` — nothing enters R memory until you `collect()`.

### Simple:

```{r}
#| eval: false

library(duckdb)
library(DBI)
con <- dbConnect(duckdb())

# Register and query a parquet file directly
dbGetQuery(con, "
  SELECT region, SUM(sales) AS total_sales
  FROM read_parquet('data/sales.parquet')
  GROUP BY region
  ORDER BY total_sales DESC
")
```

### Advanced:

```{r}
#| eval: false

library(duckdb)
library(DBI)
con <- dbConnect(duckdb())

# ── Query a single Parquet file ────────────────────────────────────────────────
dbGetQuery(
  con,
  "
  SELECT
    event_name,
    COUNT(*) AS events,
    COUNT(DISTINCT user_pseudo_id) AS users
  FROM read_parquet('data/ga4_flat.parquet')
  GROUP BY event_name
  ORDER BY events DESC
  LIMIT 10
"
)

# ── Query multiple Parquet files with a wildcard ───────────────────────────────
# Perfect for GA4's one-file-per-day structure
dbGetQuery(
  con,
  "
  SELECT
    event_date,
    COUNT(*) AS daily_events
  FROM read_parquet('data/ga4_*.parquet')
  GROUP BY event_date
  ORDER BY event_date
"
)

dbGetQuery(
  con,
  "
  SELECT year, SUM(revenue) AS total
  FROM read_parquet('data/sales/*.parquet', union_by_name = TRUE)
  GROUP BY year
  ORDER BY year
"

# ── Join two Parquet files in a single query ───────────────────────────────────
dbGetQuery(
  con,
  "
  SELECT
    s.device_category,
    s.traffic_source_medium,
    COUNT(DISTINCT e.user_pseudo_id) AS users,
    SUM(CASE WHEN e.event_name = 'purchase' THEN 1 ELSE 0 END) AS purchases
  FROM read_parquet('data/ga4_flat.parquet')    AS e
  JOIN read_parquet('data/ga4_sessions.parquet') AS s
    ON e.user_pseudo_id = s.user_pseudo_id
   AND e.event_date     = s.event_date
  GROUP BY s.device_category, s.traffic_source_medium
  ORDER BY users DESC
"
)

dbDisconnect(con, shutdown = TRUE)
```

> This is powerful — DuckDB pushes the entire query down to the file, reading only the columns and rows it needs (thanks to Parquet's columnar layout).

## Parttern 2. Use DuckDB as a dplyr backend (no SQL required)

> If you prefer dplyr verbs over SQL, you can register the **Parquet file** as a **virtual table** and query it lazily.
>
> `dbplyr` translates dplyr verbs into DuckDB SQL automatically:

### Simple:

```{r}
#| eval: false
library(duckdb)
library(dplyr)
library(dbplyr)
library(DBI)

con <- dbConnect(duckdb())

duckdb_register(con, "sales", read_parquet("data/sales.parquet"))

tbl(con, "sales") |>
  filter(year >= 2022) |>
  group_by(region) |>
  summarise(total = sum(sales)) |>
  arrange(desc(total)) |>
  collect()   # only now does data enter R
```

### Advanced:

```{r}
#| eval: false
library(duckdb)
library(dplyr)
library(dbplyr)
library(DBI)

con <- dbConnect(duckdb())

# ── Register Parquet file as a virtual table ───────────────────────────────────
# This creates a view — no data is copied or loaded
duckdb_read_csv   # (there is also this for CSV)

dbExecute(con, "
  CREATE VIEW ga4_events AS
  SELECT * FROM read_parquet('data/ga4_flat.parquet')
")

dbExecute(con, "
  CREATE VIEW ga4_sessions AS
  SELECT * FROM read_parquet('data/ga4_sessions.parquet')
")

# ── Now use dplyr as if it were a regular data frame ──────────────────────────
events   <- tbl(con, "ga4_events")
sessions <- tbl(con, "ga4_sessions")

# dplyr query — translated to DuckDB SQL by dbplyr
top_events <- events |>
  count(event_name, sort = TRUE) |>
  head(10) |>
  collect()    # collect() executes the query and returns a tibble

print(top_events)

# Join two Parquet files using dplyr — no SQL written
device_conversions <- sessions |>
  select(user_pseudo_id, device_category, traffic_source_medium) |>
  left_join(
    events |>
      filter(event_name == "purchase") |>
      distinct(user_pseudo_id) |>
      mutate(purchased = TRUE),
    by = "user_pseudo_id"
  ) |>
  mutate(purchased = coalesce(purchased, FALSE)) |>
  group_by(device_category, traffic_source_medium) |>
  summarise(
    users      = n_distinct(user_pseudo_id),
    purchasers = sum(as.integer(purchased), na.rm = TRUE),
    cvr        = purchasers / users,
    .groups    = "drop"
  ) |>
  arrange(desc(cvr)) |>
  collect()  # only now does data enter R

print(device_conversions)

dbDisconnect(con, shutdown = TRUE)

```

## Pattern 3. Query an Arrow Table with DuckDB (The Power Combination)

> Arrow and DuckDB are designed to work together. Arrow handles file I/O and in-memory transformation; DuckDB handles querying.
>
> They share data without copying, using the Apache Arrow memory format.
>
> If you already have an Arrow table in memory, DuckDB can query it directly — zero copying involved:

### Simple:

```{r}
#| eval: false
library(arrow)
library(duckdb)
library(dplyr)

# Create or load an Arrow table
arrow_table <- read_parquet("data/sales.parquet", as_data_frame = FALSE)

con <- dbConnect(duckdb())
duckdb_register_arrow(con, "sales_arrow", arrow_table)

tbl(con, "sales_arrow") |>
  filter(category == "Electronics") |>
  summarise(avg_price = mean(price)) |>
  collect()
```

### Advanced:

```{r}
#| eval: false

library(arrow)
library(duckdb)
library(dplyr)

# ── Step 1: Open Parquet as Arrow Dataset (lazy — nothing loaded yet) ──────────
ga4_arrow <- arrow::open_dataset("data/ga4_flat.parquet")

# ── Step 2: Do transformations in Arrow ───────────────────────────────────────
# Arrow is better at: filtering rows, selecting columns, type conversion
ga4_prepared <- ga4_arrow |>
  filter(event_name %in% c("purchase", "add_to_cart", "view_item")) |>
  select(event_date, event_name, user_pseudo_id) |>
  mutate(event_date = as.Date(event_date))

# ── Step 3: Hand off to DuckDB for aggregation ────────────────────────────────
# DuckDB is better at: GROUP BY, window functions, complex joins
con <- dbConnect(duckdb())

# Register the Arrow object directly — ZERO data copying
duckdb_register_arrow(con, "ga4_prepared", ga4_prepared)

# Now query it with full DuckDB SQL power
result <- dbGetQuery(con, "
  SELECT
    event_date,
    event_name,
    COUNT(DISTINCT user_pseudo_id) AS users
  FROM ga4_prepared
  GROUP BY event_date, event_name
  ORDER BY event_date, event_name
")

print(result)
dbDisconnect(con, shutdown = TRUE)

# ── The same workflow using dplyr throughout ───────────────────────────────────
library(arrow)
library(duckdb)
library(dplyr)

con <- dbConnect(duckdb::duckdb())

# Arrow dataset → register with DuckDB → query with dplyr
ga4_arrow <- arrow::open_dataset("data/ga4_flat.parquet")
duckdb::duckdb_register_arrow(con, "ga4_events", ga4_arrow)

ga4_params_arrow <- arrow::open_dataset("data/ga4_event_params.parquet")
duckdb::duckdb_register_arrow(con, "ga4_params", ga4_params_arrow)

# Now use pure dplyr — no SQL needed at all
# dbplyr translates these verbs to DuckDB SQL automatically

# Marketing analysis: page views by device and channel
page_analysis <- tbl(con, "ga4_params") |>
  filter(key == "page_location") |>
  rename(page_location = param_value) |>
  left_join(
    tbl(con, "ga4_events") |>
      select(event_timestamp, user_pseudo_id,
             device_category, traffic_source_medium),
    by = c("event_timestamp", "user_pseudo_id")
  ) |>
  mutate(
    page_path = regexp_replace(page_location, "\\?.*$", ""),
    page_path = regexp_replace(page_path, "https?://[^/]+", "")
  ) |>
  group_by(device_category, page_path) |>
  summarise(
    page_views = n(),
    users      = n_distinct(user_pseudo_id),
    .groups    = "drop"
  ) |>
  filter(page_views >= 10) |>
  arrange(desc(page_views)) |>
  head(20) |>
  collect()

print(page_analysis)
dbDisconnect(con, shutdown = TRUE)
```

## Pattern 4. Persistent DuckDB Database (Best for a Course Project)

> Write your rectangled GA4 tables once into a `.duckdb` file. Every subsequent analysis opens this one file — no BigQuery, no Parquet management, no RAM issues:

```{r}
#| eval: false

library(duckdb)
library(dplyr)
library(arrow)

# ── SETUP: Run once to create the database ────────────────────────────────────
con <- dbConnect(
  duckdb::duckdb(),
  dbdir     = "data/ga4.duckdb",
  read_only = FALSE
)

# Write tables from Parquet files — DuckDB reads and stores them
dbExecute(con, "
  CREATE TABLE ga4_events AS
  SELECT * FROM read_parquet('data/ga4_flat.parquet')
")

dbExecute(con, "
  CREATE TABLE ga4_params AS
  SELECT * FROM read_parquet('data/ga4_event_params.parquet')
")

dbExecute(con, "
  CREATE TABLE ga4_sessions AS
  SELECT * FROM read_parquet('data/ga4_sessions.parquet')
")

dbExecute(con, "
  CREATE TABLE ga4_items AS
  SELECT * FROM read_parquet('data/ga4_items.parquet')
")

# Add indexes for faster querying
dbExecute(con, "CREATE INDEX idx_event_name ON ga4_events(event_name)")
dbExecute(con, "CREATE INDEX idx_param_key  ON ga4_params(key)")
dbExecute(con, "CREATE INDEX idx_event_date ON ga4_events(event_date)")

# Confirm what was saved
cat("Tables in ga4.duckdb:\n")
dbListTables(con) |> print()

cat("\nRow counts:\n")
purrr::map_df(dbListTables(con), \(t) {
  tibble::tibble(
    table = t,
    rows  = dbGetQuery(con,
              glue::glue("SELECT COUNT(*) n FROM {t}"))$n
  )
}) |> print()

cat("\nFile size:",
    round(file.size("data/ga4.duckdb") / 1024^2, 1), "MB\n")

dbDisconnect(con, shutdown = TRUE)
```

```{r}
#| eval: false

# ── ANALYSIS: Open the database for every subsequent session ──────────────────
con <- dbConnect(
  duckdb::duckdb(),
  dbdir     = "data/ga4.duckdb",
  read_only = TRUE    # Prevents accidental writes during analysis
)

# Lazy table references — nothing in RAM yet
events   <- tbl(con, "ga4_events")
params   <- tbl(con, "ga4_params")
sessions <- tbl(con, "ga4_sessions")
items    <- tbl(con, "ga4_items")

# ── Example 1: Event counts (dplyr) ───────────────────────────────────────────
events |>
  count(event_name, sort = TRUE) |>
  collect()

# ── Example 2: Top pages (dplyr) ──────────────────────────────────────────────
params |>
  filter(key == "page_location") |>
  count(param_value, sort = TRUE) |>
  head(10) |>
  collect()

# ── Example 3: Revenue by day (SQL) ───────────────────────────────────────────
dbGetQuery(con, "
  SELECT
    event_date,
    COUNT(*)                                   AS transactions,
    SUM(ecommerce_purchase_revenue_in_usd)     AS revenue,
    AVG(ecommerce_purchase_revenue_in_usd)     AS avg_order_value
  FROM ga4_events
  WHERE event_name = 'purchase'
  GROUP BY event_date
  ORDER BY event_date
")

# ── Example 4: Funnel analysis (SQL window functions) ─────────────────────────
dbGetQuery(con, "
  WITH funnel AS (
    SELECT
      event_name,
      COUNT(DISTINCT user_pseudo_id) AS users,
      ROW_NUMBER() OVER (
        ORDER BY CASE event_name
          WHEN 'view_item'       THEN 1
          WHEN 'add_to_cart'     THEN 2
          WHEN 'begin_checkout'  THEN 3
          WHEN 'purchase'        THEN 4
        END
      ) AS stage_order
    FROM ga4_events
    WHERE event_name IN ('view_item','add_to_cart',
                         'begin_checkout','purchase')
    GROUP BY event_name
  )
  SELECT
    event_name,
    users,
    ROUND(100.0 * users /
      FIRST_VALUE(users) OVER (ORDER BY stage_order), 1) AS pct_of_top
  FROM funnel
  ORDER BY stage_order
")

dbDisconnect(con, shutdown = TRUE)
```

# How the Three Tools Form a Complete Stack

```         
# The complete GA4 analytics stack:

# ┌─────────────────────────────────────────────────────────────┐
# │                     YOUR R SESSION                          │
# │                                                             │
# │  ┌──────────┐    ┌──────────┐    ┌──────────────────────┐  │
# │  │  Arrow   │    │  dplyr   │    │        ggplot2        │  │
# │  │          │    │  dbplyr  │    │        gt             │  │
# │  │ Read &   │    │          │    │                       │  │
# │  │ Write    │    │ Verbs →  │    │  Visualise & Report   │  │
# │  │ Parquet  │    │ SQL      │    │                       │  │
# │  └────┬─────┘    └────┬─────┘    └───────────┬──────────┘  │
# │       │               │                       │             │
# │       └───────────────▼───────────────────────┘             │
# │                  ┌──────────┐                               │
# │                  │  DuckDB  │                               │
# │                  │          │                               │
# │                  │  Query   │                               │
# │                  │  Engine  │                               │
# │                  └────┬─────┘                               │
# └───────────────────────┼─────────────────────────────────────┘
#                         │
#          ┌──────────────┼──────────────┐
#          │              │              │
#    ┌─────▼──────┐ ┌─────▼──────┐ ┌───▼────────────┐
#    │  Parquet   │ │  Parquet   │ │  ga4.duckdb    │
#    │  ga4_flat  │ │  ga4_params│ │  (all tables)  │
#    └────────────┘ └────────────┘ └────────────────┘
#         DISK            DISK           DISK

# Arrow   = the loading dock (moves data efficiently)
# DuckDB  = the analyst (asks and answers questions)
# Parquet = the filing system (stores data efficiently)
```

# The Big Picture Workflow

```{r}
#| echo: false
#install.packages("DiagrammeR")
library(DiagrammeR)

grViz("
  digraph workflow {
    graph [layout = dot, rankdir = TD]
    node [shape = rectangle, style = filled]
    
    A [label = 'Parquet files on disk', fillcolor = '#C0DD97']
    B [label = 'DuckDB queries', fillcolor = '#B5D4F4']
    C [label = 'Large intermediate result?', fillcolor = '#FAC775', shape = diamond]
    D [label = 'compute()', fillcolor = '#B5D4F4']
    E [label = 'collect()', fillcolor = '#B5D4F4']
    F [label = 'Visualize - ggplot2', fillcolor = '#F4C0D1']
    G [label = 'Model - tidymodels', fillcolor = '#F4C0D1']
    
    A -> B -> C
    C -> D [label = 'Yes']
    C -> E [label = 'No']
    D -> E -> F
    E -> G
  }
")
```

**Parquet is your warehouse, Arrow is your forklift, DuckDB is your analyst.** Each does its job, and they hand off to each other seamlessly.

# `open_dataset` vs. `read_parquet` functions

## The core difference

> `open_dataset()` is for **many files or a folder** — it creates a lazy connection without loading anything into memory. `read_parquet()` is for **a single file** — it reads the file directly, by default into R memory.
>
> **The One-Sentence Description**:

```{{r}}
# read_parquet()  — loads data INTO RAM immediately, gives you a tibble
# open_dataset()  — creates a POINTER to data on disk, loads nothing yet
```

\> Everything else flows from this single difference. \> \> **What Each Function Actually Does**:

## `read_parquet()` — Eager Loading

```{r}
#| eval: false
library(arrow)

# read_parquet() works like read.csv() or readRDS()
# It reads the file immediately and returns a tibble in memory

ga4_data <- arrow::read_parquet("data/ga4_flat.parquet")

# After this line:
# ✓ All data is in RAM
# ✓ ga4_data is a regular tibble
# ✓ You can use it without Arrow at all — it's just a data frame
# ✗ If the file is larger than RAM — this fails or crashes R

class(ga4_data)
#> [1] "tbl_df"     "tbl"        "data.frame"
# ← a regular tibble, no Arrow in sight after loading

object.size(ga4_data) |> format(units = "MB")
#> "187.4 MB"   ← sitting in your RAM right now
```

## `open_dataset()` — Lazy Loading

```{r}
#| eval: false
library(arrow)

# open_dataset() works like a database connection
# It reads the file METADATA only and returns an Arrow Dataset object
# No actual data rows are loaded until you explicitly ask for them

ga4_dataset <- arrow::open_dataset("data/ga4_flat.parquet")

# After this line:
# ✓ Metadata is read (schema, number of rows, column types)
# ✓ You can query it immediately with dplyr verbs
# ✓ Works on files larger than RAM
# ✗ ga4_dataset is NOT a tibble — it is an Arrow Dataset object

class(ga4_dataset)
#> [1] "FileSystemDataset" "Dataset" "ArrowObject" "R6"
# ← an Arrow Dataset — NOT a data frame yet

object.size(ga4_dataset) |> format(units = "KB")
#> "4.2 KB"    ← almost nothing in RAM — just the metadata pointer
```

## Side-by-Side Comparison

```{r}
#| eval: false
library(arrow)
library(dplyr)

# ── Same question asked both ways ─────────────────────────────────────────────
# "What are the top 5 event types by count?"

# ── read_parquet() approach ───────────────────────────────────────────────────
# Step 1: Load EVERYTHING into RAM (even columns we don't need)
ga4_tibble <- read_parquet("data/ga4_flat.parquet")
# Step 2: Then filter in R
ga4_tibble |>
  count(event_name, sort = TRUE) |>
  slice_head(n = 5)

# ── open_dataset() approach ───────────────────────────────────────────────────
# Step 1: Open a pointer (nothing loaded)
ga4_ds <- open_dataset("data/ga4_flat.parquet")
# Step 2: Build a query (still nothing loaded)
query <- ga4_ds |>
  count(event_name, sort = TRUE) |>
  slice_head(n = 5)
# Step 3: Execute — Arrow reads ONLY the event_name column
query |> collect()

# Both produce identical results
# But open_dataset() used a fraction of the RAM
```

## The Key Concept: Lazy vs Eager Evaluation

```{r}
#| eval: false
library(arrow)
library(dplyr)

ga4_ds <- open_dataset("data/ga4_flat.parquet")

# ── Every dplyr verb on an Arrow Dataset is LAZY ─────────────────────────────
# Nothing happens when you call these — they just build a query plan

step1 <- ga4_ds |> filter(event_name == "purchase")
step2 <- step1  |> select(event_date, user_pseudo_id,
                            ecommerce_purchase_revenue_in_usd)
step3 <- step2  |> group_by(event_date)
step4 <- step3  |> summarise(revenue = sum(ecommerce_purchase_revenue_in_usd,
                                            na.rm = TRUE))

# At this point: ZERO rows have been read from disk
# step4 is just a description of what to do

# ── collect() EXECUTES the query ─────────────────────────────────────────────
# Only now does Arrow read from disk — and only the columns it needs
result <- step4 |> collect()

# Arrow read: event_name, event_date, user_pseudo_id,
#             ecommerce_purchase_revenue_in_usd
# Arrow skipped: every other column in the file
# This is called "column pruning" — a massive RAM saving

# ── You can inspect the query plan before executing ───────────────────────────
ga4_ds |>
  filter(event_name == "purchase") |>
  group_by(event_date) |>
  summarise(revenue = sum(ecommerce_purchase_revenue_in_usd,
                          na.rm = TRUE)) |>
  explain()    # Shows Arrow's execution plan — no data read yet
```

## The Biggest Difference: Multiple Files

This is where `open_dataset()` becomes **dramatically** more powerful.
`read_parquet()` reads one file. `open_dataset()` reads many files as
if they were one:

```{r}
#| eval: false
library(arrow)

# ── Scenario: GA4 data saved as one Parquet file per day ─────────────────────
# data/
#   ga4_20201101.parquet
#   ga4_20201102.parquet
#   ga4_20201103.parquet
#   ... (365 files)

# ── read_parquet(): must read files one at a time ─────────────────────────────
library(purrr)

# Cumbersome — load all files and bind them
all_files <- list.files("data", pattern = "ga4_\\d{8}.parquet",
                         full.names = TRUE)

ga4_all <- all_files |>
  map(\(f) read_parquet(f)) |>
  list_rbind()
# Loads ALL files into RAM simultaneously
# If 365 files × 50 MB each = 18 GB — crashes R

# ── open_dataset(): treats ALL files as one dataset ──────────────────────────
ga4_all <- open_dataset("data/")   # Point at the folder, not a file
# OR be more specific:
ga4_all <- open_dataset(
  sources = list.files("data", pattern = "ga4_\\d{8}.parquet",
                        full.names = TRUE)
)

# Now query across ALL 365 days as if it were one table
# Arrow reads only the files and columns needed to answer your query
ga4_all |>
  filter(event_date >= "2021-01-01",
         event_date <= "2021-01-31") |>
  count(event_name) |>
  collect()
# Arrow only opens January files — skips the other 334 files entirely
```

## Hive-Partitioned Datasets — The Most Powerful Pattern

```{r}
#| eval: false
library(arrow)
library(dplyr)

# ── Save GA4 data in Hive partition format ────────────────────────────────────
# Hive partitioning stores data in folders named key=value
# This lets open_dataset() skip entire folders based on filter conditions

ga4_tibble <- readRDS("data/ga4_raw.rds") |>
  mutate(
    year  = lubridate::year(lubridate::ymd(event_date)),
    month = lubridate::month(lubridate::ymd(event_date))
  )

# Write partitioned dataset
arrow::write_dataset(
  ga4_tibble,
  path             = "data/ga4_partitioned",
  format           = "parquet",
  partitioning     = c("year", "month"),   # Partition by year and month
  compression      = "snappy"
)

# Creates this folder structure:
# data/ga4_partitioned/
#   year=2020/
#     month=11/
#       part-0.parquet
#     month=12/
#       part-0.parquet
#   year=2021/
#     month=1/
#       part-0.parquet
#     month=2/
#       part-0.parquet

# ── open_dataset() understands Hive partitioning automatically ────────────────
ga4_partitioned <- open_dataset(
  "data/ga4_partitioned",
  partitioning = c("year", "month")   # Tell Arrow how it is partitioned
)

# Query January 2021 — Arrow reads ONLY year=2021/month=1/ folder
jan_2021 <- ga4_partitioned |>
  filter(year == 2021, month == 1) |>
  count(event_name) |>
  collect()

# Arrow skipped every other year= and month= folder
# This is called "partition pruning" — can be 100x faster on large datasets

# ── Verify the schema includes partition columns ──────────────────────────────
ga4_partitioned$schema
#> Schema
#> event_date: string
#> event_name: string
#> user_pseudo_id: string
#> ...
#> year: int32      ← added automatically from folder name
#> month: int32     ← added automatically from folder name
```

## Comparison table

|   | `open_dataset()` | `read_parquet()` |
|------------------------|------------------------|------------------------|
| Input | Folder, multiple files, or single file | Single file only |
| Loads into memory? | No — lazy connection | Yes by default |
| Returns | Arrow Dataset object | R dataframe or Arrow Table |
| Partitioned folders? | Yes, auto-detects partitions | No |
| Multiple file formats? | Yes (Parquet, CSV, Arrow) | Parquet only |
| Use with dplyr? | Yes, lazily | Yes, but already in memory |
| Best for | Large, partitioned, multi-file data | Small-to-medium single files |


### Detailed Feature Comparison Table

```{r}
#| echo: false
tibble::tribble(
  ~Feature,                        ~read_parquet,              ~open_dataset,
  "Returns",                       "tibble (in RAM)",           "Arrow Dataset (pointer)",
  "Data loaded immediately",       "Yes — all of it",           "No — metadata only",
  "Works on single file",          "Yes",                       "Yes",
  "Works on multiple files",       "One at a time",             "Yes — all as one table",
  "Works on a folder",             "No",                        "Yes",
  "Hive partitioning support",     "No",                        "Yes",
  "Column pruning",                "No — loads all columns",    "Yes — loads only needed columns",
  "Partition pruning",             "No",                        "Yes — skips irrelevant files",
  "Larger-than-RAM files",         "No — will crash",           "Yes — streams from disk",
  "dplyr verbs work",              "Yes — immediately",         "Yes — lazily until collect()",
  "collect() required",            "No — already a tibble",     "Yes — to get a tibble",
  "Schema inspection",             "After loading",             "Before loading",
  "Compression formats",           "All Arrow supports",        "All Arrow supports",
  "Best for",                      "Small-medium files",        "Large or multiple files"
) |>
  gt::gt() |>
  gt::tab_header(
    title    = "read_parquet() vs open_dataset()",
    subtitle = "Complete feature comparison"
  ) |>
  gt::cols_label(
    Feature       = "Feature",
    read_parquet  = "read_parquet()",
    open_dataset  = "open_dataset()"
  ) |>
  gt::tab_style(
    style     = gt::cell_fill(color = "#e8f5e9"),
    locations = gt::cells_body(
      rows = Feature %in% c(
        "Larger-than-RAM files",
        "Works on multiple files",
        "Hive partitioning support",
        "Column pruning",
        "Partition pruning"
      )
    )
  ) |>
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::opt_stylize(style = 1)
```

## Similarities — What They Share

```{r}
#| eval: false
library(arrow)
library(dplyr)

# ── Both read Parquet format ──────────────────────────────────────────────────
# Both understand snappy, gzip, zstd, lz4 compression
# Both read the same column types: string, int32, float64, bool, date, etc.

# ── Both work with dplyr verbs ────────────────────────────────────────────────
# (open_dataset lazily, read_parquet eagerly)
result_rp <- read_parquet("data/ga4_flat.parquet") |>
  filter(event_name == "purchase") |>
  count(event_date)

result_od <- open_dataset("data/ga4_flat.parquet") |>
  filter(event_name == "purchase") |>
  count(event_date) |>
  collect()

# Results are identical
identical(result_rp, result_od)
#> TRUE

# ── Both support column selection ─────────────────────────────────────────────
# read Parquet — use col_select to avoid loading unused columns
ga4_small_rp <- read_parquet(
  "data/ga4_flat.parquet",
  col_select = c("event_date", "event_name", "user_pseudo_id")
)

# open_dataset — use select() lazily
ga4_small_od <- open_dataset("data/ga4_flat.parquet") |>
  select(event_date, event_name, user_pseudo_id) |>
  collect()

# ── Both can read multiple formats ────────────────────────────────────────────
# Parquet, Arrow IPC (feather), CSV, JSON, ORC

# read_parquet is Parquet-specific, but open_dataset handles all:
open_dataset("data/ga4.csv",    format = "csv")
open_dataset("data/ga4.feather", format = "feather")
open_dataset("data/ga4.parquet", format = "parquet")  # default

# ── Both support schema inspection ────────────────────────────────────────────
# read_parquet schema (after loading)
read_parquet("data/ga4_flat.parquet") |> glimpse()

# open_dataset schema (before loading anything)
open_dataset("data/ga4_flat.parquet")$schema
#> Schema
#> event_date: string
#> event_name: string
#> user_pseudo_id: string
#> device_category: string
#> ...
```

## Practical GA4 Workflows

### Workflow 1: `read_parquet()` for Quick Exploration

```{r}
#| eval: false
library(arrow)
library(dplyr)

# You just created a small processed file and want to inspect it
# File is 15 MB — perfectly fine to load into RAM

ga4_items <- read_parquet("data/ga4_items.parquet")

# Now it is just a tibble — use any R function freely
summary(ga4_items$price_in_usd)
hist(ga4_items$quantity)
View(ga4_items)

# No collect() needed anywhere
ga4_items |>
  filter(item_category == "Apparel") |>
  arrange(desc(price_in_usd)) |>
  head(10)
```

### Workflow 2: `open_dataset()` for Large Event Params

```{r}
#| eval: false
library(arrow)
library(dplyr)

# event_params unrolled is very large — open_dataset is safer
ga4_params <- open_dataset("data/ga4_event_params.parquet")

# Check size without loading
ga4_params$schema          # Column names and types
nrow(ga4_params)           # Row count — reads metadata only, instant

# Query only what you need — Arrow loads only 2 columns from disk
page_views <- ga4_params |>
  filter(
    event_name  == "page_view",
    key         == "page_location"
  ) |>
  select(event_date, user_pseudo_id, param_value) |>
  collect()   # Only now does data enter RAM — just the filtered subset

nrow(page_views)   # Much smaller than the full params file
```

### Workflow 3: `open_dataset()` Across All GA4 Daily Files

```{r}
#| eval: false
library(arrow)
library(dplyr)
library(lubridate)

# GA4 exports one file per day — open_dataset handles all of them
# Save each day as its own Parquet file first:

# data/ga4_daily/
#   ga4_20201101.parquet
#   ga4_20201102.parquet
#   ... etc

# Open all daily files as one unified dataset
ga4_all <- open_dataset(
  sources = list.files(
    "data/ga4_daily",
    pattern    = "\\.parquet$",
    full.names = TRUE
  )
)

cat("Total rows across all files:", nrow(ga4_all), "\n")
cat("Schema:\n")
print(ga4_all$schema)

# Query across the full date range — Arrow only reads what is needed
monthly_summary <- ga4_all |>
  mutate(
    event_date = as.Date(event_date),
    month      = floor_date(event_date, "month")
  ) |>
  group_by(month, event_name) |>
  summarise(
    events = n(),
    users  = n_distinct(user_pseudo_id),
    .groups = "drop"
  ) |>
  collect()

monthly_summary |>
  filter(event_name %in% c("page_view", "purchase",
                             "add_to_cart")) |>
  ggplot2::ggplot(
    ggplot2::aes(x = month, y = events,
                  colour = event_name,
                  group  = event_name)
  ) +
  ggplot2::geom_line(linewidth = 1.1) +
  ggplot2::geom_point(size = 2.5) +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::scale_x_date(date_labels = "%b %Y") +
  ggplot2::labs(
    title  = "Monthly Event Trends",
    x      = NULL,
    y      = "Events",
    colour = "Event Type"
  ) +
  ggplot2::theme_minimal(base_size = 13)
```

### Workflow 4: Combining Both in One Pipeline

```{r}
#| eval: false
library(arrow)
library(dplyr)

# Common pattern: open_dataset for the large file, read_parquet for small ones
# Then join them

# Large file — use open_dataset (lazy)
ga4_params <- open_dataset("data/ga4_event_params.parquet")

# Small lookup file — use read_parquet (eager, it's tiny)
page_categories <- read_parquet("data/page_categories.parquet")
# This is a small reference table: page_path → category
# Fine to have in RAM as a regular tibble

# Extract page locations from the large params file lazily
page_data <- ga4_params |>
  filter(key == "page_location") |>
  select(event_date, user_pseudo_id, param_value) |>
  rename(page_url = param_value) |>
  mutate(
    page_path = arrow_string_replace_regex(
      page_url,
      "^https?://[^/]+|\\?.*$", ""
    )
  ) |>
  collect()   # Now bring into RAM — filtered to just page views

# Join with the small lookup table (both in RAM now)
page_data |>
  left_join(page_categories, by = "page_path") |>
  count(category, sort = TRUE)
```

### Workflow 5: `open_dataset()` with DuckDB — The Full Stack

```{r}
#| eval: false
library(arrow)
library(duckdb)
library(dplyr)

# open_dataset creates Arrow Dataset objects
# DuckDB can query Arrow objects directly — zero data copying

ga4_events_ds <- open_dataset("data/ga4_flat.parquet")
ga4_params_ds <- open_dataset("data/ga4_event_params.parquet")

# Register Arrow datasets with DuckDB
con <- dbConnect(duckdb::duckdb())
duckdb::duckdb_register_arrow(con, "events", ga4_events_ds)
duckdb::duckdb_register_arrow(con, "params", ga4_params_ds)

# Now use DuckDB SQL on top of Arrow Datasets
# Data stays in Arrow format — no copying between systems
DBI::dbGetQuery(con, "
  SELECT
    e.event_date,
    e.device_category,
    COUNT(DISTINCT e.user_pseudo_id)           AS users,
    SUM(e.ecommerce_purchase_revenue_in_usd)   AS revenue
  FROM events e
  WHERE e.event_name = 'purchase'
  GROUP BY e.event_date, e.device_category
  ORDER BY e.event_date, revenue DESC
")

DBI::dbDisconnect(con, shutdown = TRUE)
```

## Which to use for GA4 data?


```{{r}}
# Ask these questions in order:

# 1. How big is the file?
#    < 500 MB and fits in RAM  → read_parquet() is fine
#    > 500 MB or close to RAM  → open_dataset()

# 2. How many files?
#    One file                  → either works
#    Multiple files or folder  → open_dataset() only

# 3. How many columns do you need?
#    Most or all columns       → read_parquet() is simpler
#    Only a few columns        → open_dataset() (column pruning saves RAM)

# 4. Are you exploring or producing?
#    Exploring unknown data    → read_parquet() (View(), summary(), hist())
#    Building analysis pipeline → open_dataset() (lazy, memory-efficient)

# 5. Do you need the result as a tibble immediately?
#    Yes                       → read_parquet()
#    No — you will filter first → open_dataset() + collect()
```