---
title: "Duckplyr: The Best of Both Worlds"
subtitle: "drop-in replacement for dplyr"
description: "This page describes the benifits of using duckplyr and illustrate how to use it for GA4 data"
categories: [Duckplyr, DuckDB, dplyr]
order: 3
author: "Jae Jung"
date: today
format: 
    html:
        theme: cosmo
        toc: true
        toc-depth: 3
        toc-expand: 1
        code-fold: show
        toc-location: right-body
        number-sections: true
        #embed-resources: true
execute:
  warning: false
  message: false
  echo: fenced
  cache: false
  freeze: auto
---

# **What is duckplyr?**

> duckplyr wraps DuckDB's analytical query processing engine in a dplyr-compatible API. In plain terms: it's **dplyr, but secretly powered by DuckDB under the hood.** Calling `library(duckplyr)` overwrites dplyr methods, enabling duckplyr for the entire session. Your existing dplyr code runs unchanged — you just get DuckDB's speed for free.
>
> Duckplyr is a **drop-in replacement for dplyr that uses DuckDB as its computational engine**. It was created by the same team that built DuckDB and is designed to make DuckDB's speed available through the exact same interface you already use for dplyr — with zero syntax changes.

```{r}
#| eval: false
# The elevator pitch:
# dplyr   = friendly syntax, moderate speed, RAM-limited
# DuckDB  = fast engine, requires SQL or dbplyr setup, RAM-independent
# duckplyr = friendly syntax + fast engine + RAM-independent
#            ↑ you get all three at once
```

> The key insight is that duckplyr is not a wrapper around dplyr or a translation layer like dbplyr. It **reimplements dplyr's verbs** using DuckDB's execution engine directly, so the familiar code you write is executed by DuckDB under the hood — automatically, invisibly.

# **The Landscape: Four Tools, One Job**

Before comparing them, let us be precise about what each tool actually is:

```{r}
#| eval: false
# All four answer the question: "how do I manipulate tabular data in R?"
# They differ in WHERE computation happens and HOW you write the code

# ── dplyr ─────────────────────────────────────────────────────────────────────
# Computation: R's in-memory engine
# Interface:   dplyr verbs
# Data must:   fit in RAM
# Speed:       moderate

library(dplyr)
ga4_events |>
  filter(event_name == "purchase") |>
  group_by(event_date) |>
  summarise(revenue = sum(revenue))

# ── DuckDB (via DBI) ──────────────────────────────────────────────────────────
# Computation: DuckDB engine
# Interface:   SQL strings
# Data must:   NOT need to fit in RAM
# Speed:       very fast

library(DBI)
library(duckdb)
dbGetQuery(
  con,
  "
  SELECT event_date, SUM(revenue) AS revenue
  FROM ga4_events
  WHERE event_name = 'purchase'
  GROUP BY event_date
"
)

# ── DuckDB (via dbplyr) ───────────────────────────────────────────────────────
# Computation: DuckDB engine
# Interface:   dplyr verbs (translated to SQL by dbplyr)
# Data must:   NOT need to fit in RAM
# Speed:       very fast
# Catch:       not all dplyr verbs translate perfectly

library(dbplyr)
tbl(con, "ga4_events") |>
  filter(event_name == "purchase") |>
  group_by(event_date) |>
  summarise(revenue = sum(revenue)) |>
  collect()

# ── duckplyr ─────────────────────────────────────────────────────────────────
# Computation: DuckDB engine
# Interface:   dplyr verbs (NATIVE — not translated)
# Data must:   NOT need to fit in RAM
# Speed:       very fast
# Catch:       falls back to dplyr when needed (gracefully)

library(duckplyr)
ga4_events |> # ← same code as dplyr
  filter(event_name == "purchase") |> # ← no con object needed
  group_by(event_date) |> # ← no collect() needed
  summarise(revenue = sum(revenue)) # ← result is a tibble automatically
```

# **How duckplyr Is Different from Each Tool**

## duckplyr vs dplyr

```{r}
#| eval: false
# ── What changes: ─────────────────────────────────────────────────────────────
#   library(dplyr)    →   library(duckplyr)
#   That's it. Seriously.

# ── What stays the same: ──────────────────────────────────────────────────────
#   All verb names: filter, select, mutate, group_by, summarise,
#                   arrange, join, slice, rename, distinct...
#   All syntax: pipes, quasiquotation, across(), where()...
#   All output: tibbles, same column names, same data types

# ── What is different under the hood: ─────────────────────────────────────────
#   dplyr    executes in R's vector engine
#   duckplyr executes in DuckDB's columnar engine

# ── Performance difference: ───────────────────────────────────────────────────
# For small data (< 100k rows): negligible, sometimes dplyr is faster
# For large data (> 1M rows):   duckplyr is 10-100x faster
# For larger-than-RAM data:     dplyr fails, duckplyr works
```

## duckplyr vs DuckDB + dbplyr

```{r}
#| eval: false
# ── DuckDB + dbplyr requires: ─────────────────────────────────────────────────
# 1. Create a connection object
con <- dbConnect(duckdb::duckdb(), "data/ga4.duckdb")
# 2. Create table references
events <- tbl(con, "ga4_events")
# 3. Write your query
result <- events |>
  filter(...) |>
  summarise(...) |>
  # 4. Explicitly collect results
  collect()
# 5. Remember to disconnect
dbDisconnect(con, shutdown = TRUE)

# ── duckplyr requires: ────────────────────────────────────────────────────────
# 1. Load the library
library(duckplyr)
# 2. Write your query (same as dplyr)
result <- ga4_events |> filter(...) |> summarise(...)
# That's it — no connection, no collect(), no disconnect

# ── The translation gap: ──────────────────────────────────────────────────────
# dbplyr TRANSLATES dplyr to SQL — some verbs have no SQL equivalent
# duckplyr REIMPLEMENTS dplyr using DuckDB — much more complete coverage

# Example: dbplyr cannot translate some R functions
tbl(con, "ga4_events") |>
  mutate(path = stringr::str_remove(page_url, "\\?.*$")) |>
  collect()
#> Error: str_remove() is not supported in SQL translation

# duckplyr handles it natively
ga4_events |>
  mutate(path = stringr::str_remove(page_url, "\\?.*$"))
#> Works — DuckDB handles the regex directly
```

## The Fallback System — Why duckplyr Is Safe to Use

This is duckplyr's most important design feature. When it encounters something it cannot execute in DuckDB, it **silently falls back to dplyr** rather than throwing an error:

```{r}
#| eval: false
library(duckplyr)

# duckplyr executes this in DuckDB (fast path)
result1 <- ga4_events |>
  filter(event_name == "purchase") |>
  group_by(event_date) |>
  summarise(n = n())
# → executed by DuckDB engine ✓

# duckplyr cannot handle this custom R function in DuckDB
my_custom_function <- function(x) paste0("[", x, "]")

result2 <- ga4_events |>
  mutate(labeled = my_custom_function(event_name))
# → falls back to dplyr silently ✓ (no error, correct result)

# You can monitor fallbacks to understand your code
duckplyr::fallback_sitrep()
# Reports which operations fell back and why
```

```{r}
#| eval: false
# Control fallback behaviour explicitly
options(duckplyr.fallback = "warn") # Warn when falling back
options(duckplyr.fallback = "error") # Error instead of falling back
options(duckplyr.fallback = "ignore") # Silent fallback (default)
```

# **Workfolow Guide**

## Installing and Setting Up duckplyr

```{r}
#| eval: false

# ── Install ───────────────────────────────────────────────────────────────────
install.packages("duckplyr")

# ── Two ways to use it ────────────────────────────────────────────────────────

# Option A: Load alongside dplyr (duckplyr methods take precedence)
library(dplyr)
library(duckplyr)
# Now dplyr verbs on data frames use DuckDB automatically

# Option B: Use as a complete dplyr replacement
library(duckplyr)
# Masks dplyr functions — everything goes through DuckDB
# This is the recommended approach

# ── Check version ─────────────────────────────────────────────────────────────
packageVersion("duckplyr") # Should be >= 0.4.0

# ── Opt in to telemetry or out ────────────────────────────────────────────────
duckplyr::duckplyr_telemetry_set(FALSE) # Disable usage reporting
```

## Reading Data into duckplyr

duckplyr has native methods for reading Parquet and CSV files directly:

```{r}
#| eval: false
library(duckplyr)
library(arrow)

# ── Method 1: From a tibble already in memory ─────────────────────────────────
ga4_tibble <- readRDS("data/ga4_raw.rds")

# Convert tibble to duckplyr frame (enables DuckDB execution)
ga4_duck <- as_duckplyr_df(ga4_tibble)
class(ga4_duck)
#> [1] "duckplyr_df" "tbl_df" "tbl" "data.frame"

# ── Method 2: Read Parquet directly — most powerful approach ──────────────────
# No RAM used — DuckDB streams from disk as needed
ga4_events <- duckplyr::df_from_parquet("data/ga4_flat.parquet")
ga4_params <- duckplyr::df_from_parquet("data/ga4_event_params.parquet")
ga4_sessions <- duckplyr::df_from_parquet("data/ga4_sessions.parquet")
ga4_items <- duckplyr::df_from_parquet("data/ga4_items.parquet")

# Each of these is a duckplyr_df — a lazy reference to the Parquet file
# Nothing is loaded into RAM until you need actual values
class(ga4_events)
#> [1] "duckplyr_df" "tbl_df" "tbl" "data.frame"

# ── Method 3: Read multiple Parquet files with wildcard ───────────────────────
# Perfect for GA4's one-file-per-day pattern
ga4_all_days <- duckplyr::df_from_parquet("data/ga4_day_*.parquet")

# ── Method 4: From Arrow dataset ──────────────────────────────────────────────
ga4_arrow <- arrow::open_dataset("data/ga4_flat.parquet")
ga4_duck <- as_duckplyr_df(ga4_arrow)
```

# **Analyzing GA4 Data with duckplyr**

Now the practical part. Every analysis below uses **identical syntax to dplyr** but executes in DuckDB.

## Setup

```{r}
#| eval: false
library(duckplyr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(gt)
library(glue)
library(lubridate)
library(stringr)
library(forcats)

# Load all GA4 tables as duckplyr frames (lazy — nothing in RAM yet)
ga4_events <- duckplyr::df_from_parquet("data/ga4_flat.parquet")
ga4_params <- duckplyr::df_from_parquet("data/ga4_event_params.parquet")
ga4_sessions <- duckplyr::df_from_parquet("data/ga4_sessions.parquet")
ga4_items <- duckplyr::df_from_parquet("data/ga4_items.parquet")

# Confirm they are duckplyr frames
cat("ga4_events class:  ", class(ga4_events)[1], "\n")
cat("ga4_params class:  ", class(ga4_params)[1], "\n")

# Peek at available columns
glimpse(ga4_events)
glimpse(ga4_params)

```

## Analysis 1: Event Volume Overview

```{r}
#| eval: false

# ── Identical to dplyr — but DuckDB executes it ───────────────────────────────
event_summary <- ga4_events |>
  group_by(event_name) |>
  summarise(
    total_events = n(),
    unique_users = n_distinct(user_pseudo_id),
    unique_days = n_distinct(event_date),
    events_per_day = round(total_events / unique_days, 1),
    .groups = "drop"
  ) |>
  arrange(desc(total_events))

# Result is already a tibble — no collect() needed
event_summary |>
  gt() |>
  tab_header(
    title = "GA4 Event Summary",
    subtitle = "Computed by DuckDB via duckplyr"
  ) |>
  fmt_number(columns = c(total_events, unique_users), decimals = 0) |>
  fmt_number(columns = events_per_day, decimals = 1) |>
  cols_label(
    event_name = "Event",
    total_events = "Total Events",
    unique_users = "Unique Users",
    unique_days = "Days",
    events_per_day = "Events/Day"
  ) |>
  opt_stylize(style = 1)
```

## Analysis 2: Traffic Source Performance

```{r}
#| eval: false

# ── Join sessions with purchase events ────────────────────────────────────────
# duckplyr handles joins the same as dplyr
purchasers <- ga4_events |>
  filter(event_name == "purchase") |>
  distinct(user_pseudo_id) |>
  mutate(purchased = TRUE)

traffic_performance <- ga4_sessions |>
  mutate(
    channel = case_when(
      is.na(traffic_source_medium) |
        traffic_source_medium == "(none)" ~ "Direct",
      traffic_source_medium == "organic" ~ "Organic Search",
      traffic_source_medium %in% c("cpc", "ppc") ~ "Paid Search",
      str_detect(traffic_source_medium, "email") ~ "Email",
      str_detect(traffic_source_medium, "social") ~ "Social",
      TRUE ~ str_to_title(traffic_source_medium)
    )
  ) |>
  group_by(channel) |>
  summarise(
    sessions = n(),
    unique_users = n_distinct(user_pseudo_id),
    .groups = "drop"
  ) |>
  left_join(
    ga4_sessions |>
      mutate(
        channel = case_when(
          is.na(traffic_source_medium) |
            traffic_source_medium == "(none)" ~ "Direct",
          traffic_source_medium == "organic" ~ "Organic Search",
          traffic_source_medium %in% c("cpc", "ppc") ~ "Paid Search",
          str_detect(traffic_source_medium, "email") ~ "Email",
          str_detect(traffic_source_medium, "social") ~ "Social",
          TRUE ~ str_to_title(traffic_source_medium)
        )
      ) |>
      select(user_pseudo_id, channel) |>
      left_join(purchasers, by = "user_pseudo_id") |>
      group_by(channel) |>
      summarise(
        converters = sum(as.integer(!is.na(purchased)), na.rm = TRUE),
        .groups = "drop"
      ),
    by = "channel"
  ) |>
  mutate(
    cvr = converters / unique_users,
    sessions_pct = sessions / sum(sessions)
  ) |>
  arrange(desc(sessions))

traffic_performance |>
  gt() |>
  tab_header(title = "Traffic Channel Performance") |>
  fmt_number(columns = c(sessions, unique_users, converters), decimals = 0) |>
  fmt_percent(columns = c(cvr, sessions_pct), decimals = 1) |>
  cols_label(
    channel = "Channel",
    sessions = "Sessions",
    unique_users = "Users",
    converters = "Buyers",
    cvr = "CVR",
    sessions_pct = "% of Sessions"
  ) |>
  data_color(
    columns = cvr,
    method = "numeric",
    palette = c("#ffffff", "#43a04740")
  ) |>
  opt_stylize(style = 1)
```

## Analysis 3: Page Performance from `event_params`

```{r}
#| eval: false

# ── Work directly with the long-format params table ───────────────────────────
# duckplyr reads the Parquet file and filters on disk — no RAM spike

top_pages <- ga4_params |>
  filter(
    event_name == "page_view",
    key == "page_location",
    !is.na(param_value)
  ) |>
  mutate(
    # Parse URL path — duckplyr passes regex to DuckDB
    page_path = str_remove(param_value, "^https?://[^/]+"),
    page_path = str_remove(page_path, "\\?.*$"),
    page_path = str_remove(page_path, "#.*$"),
    page_path = if_else(page_path == "" | is.na(page_path), "/", page_path),
    # Categorise by section
    section = case_when(
      page_path == "/" ~ "Home",
      str_detect(page_path, "^/google\\+re|/store") ~ "Store",
      str_detect(page_path, "^/product") ~ "Products",
      str_detect(page_path, "^/basket|/cart") ~ "Cart",
      str_detect(page_path, "^/checkout") ~ "Checkout",
      TRUE ~ "Other"
    )
  ) |>
  group_by(section, page_path) |>
  summarise(
    page_views = n(),
    unique_users = n_distinct(user_pseudo_id),
    .groups = "drop"
  ) |>
  arrange(desc(page_views)) |>
  slice_head(n = 20)

# Visualise
top_pages |>
  slice_head(n = 15) |>
  mutate(
    page_path = str_trunc(page_path, 40),
    page_path = fct_reorder(page_path, page_views)
  ) |>
  ggplot(aes(x = page_path, y = page_views, fill = section)) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = comma(page_views)),
    hjust = -0.15,
    size = 3.5
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = comma,
    expand = expansion(mult = c(0, 0.18))
  ) +
  scale_fill_brewer(palette = "Set2", name = "Section") +
  labs(
    title = "Top 15 Pages by Page Views",
    subtitle = "Computed via duckplyr on Parquet — no full load into RAM",
    x = NULL,
    y = "Page Views"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## Analysis 4: Purchase Funnel

```{r}
#| eval: false

# ── Funnel stages ─────────────────────────────────────────────────────────────
funnel_stages <- c(
  "view_item",
  "add_to_cart",
  "begin_checkout",
  "add_payment_info",
  "purchase"
)

funnel_labels <- c(
  "view_item" = "View Item",
  "add_to_cart" = "Add to Cart",
  "begin_checkout" = "Begin Checkout",
  "add_payment_info" = "Add Payment",
  "purchase" = "Purchase"
)

# ── Count users at each stage using duckplyr ──────────────────────────────────
funnel_data <- ga4_events |>
  filter(event_name %in% funnel_stages) |>
  group_by(event_name) |>
  summarise(
    users = n_distinct(user_pseudo_id),
    .groups = "drop"
  ) |>
  mutate(
    stage_label = recode(event_name, !!!funnel_labels),
    stage_order = match(event_name, funnel_stages)
  ) |>
  arrange(stage_order) |>
  mutate(
    pct_of_top = users / max(users),
    pct_prev = users / lag(users),
    drop_off = 1 - pct_prev
  )

# ── Visualise ─────────────────────────────────────────────────────────────────
funnel_data |>
  mutate(stage_label = fct_inorder(stage_label)) |>
  ggplot(aes(x = stage_label, y = users, fill = pct_of_top)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(
    aes(label = glue("{comma(users)}\n{percent(pct_of_top, accuracy=0.1)}")),
    vjust = -0.4,
    size = 3.8,
    lineheight = 0.85
  ) +
  geom_text(
    data = funnel_data |> filter(!is.na(drop_off)),
    aes(
      x = stage_order - 0.5,
      y = max(funnel_data$users) * 0.6,
      label = glue("▼{percent(drop_off, accuracy=1)}")
    ),
    colour = "#e53935",
    size = 3.5
  ) +
  scale_fill_gradient(low = "#bbdefb", high = "#1565c0") +
  scale_y_continuous(
    labels = comma,
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title = "Purchase Funnel",
    subtitle = "duckplyr: DuckDB engine, dplyr syntax",
    x = NULL,
    y = "Unique Users"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(size = 11))
```

## Analysis 5: E-commerce Revenue Analysis

```{r}
#| eval: false

# ── Daily revenue with 7-day rolling average ──────────────────────────────────
# duckplyr supports window functions natively
daily_revenue <- ga4_events |>
  filter(event_name == "purchase") |>
  group_by(event_date) |>
  summarise(
    transactions = n(),
    revenue = sum(ecommerce_purchase_revenue_in_usd, na.rm = TRUE),
    aov = mean(ecommerce_purchase_revenue_in_usd, na.rm = TRUE),
    unique_buyers = n_distinct(user_pseudo_id),
    .groups = "drop"
  ) |>
  arrange(event_date) |>
  mutate(
    event_date = as.Date(event_date),
    # 7-day rolling average — duckplyr handles this via DuckDB window functions
    revenue_7day = slider::slide_dbl(
      revenue,
      mean,
      .before = 6,
      .complete = FALSE
    )
  )

# ── Revenue chart ─────────────────────────────────────────────────────────────
daily_revenue |>
  ggplot(aes(x = event_date)) +
  geom_col(aes(y = revenue), fill = "#90caf9", width = 0.8, alpha = 0.7) +
  geom_line(aes(y = revenue_7day), colour = "#1565c0", linewidth = 1.2) +
  geom_point(aes(y = revenue_7day), colour = "#1565c0", size = 2) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week") +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Daily Revenue with 7-Day Rolling Average",
    subtitle = "Bars = daily revenue · Line = 7-day rolling average",
    x = NULL,
    y = "Revenue (USD)"
  ) +
  theme_minimal(base_size = 13)
```

## Analysis 6: Product Performance (from `ga4_items`)

```{r}
#| eval: false

# ── Top products and categories ────────────────────────────────────────────────
product_performance <- ga4_items |>
  filter(!is.na(item_name)) |>
  group_by(item_category, item_name) |>
  summarise(
    units_sold = sum(quantity, na.rm = TRUE),
    revenue = sum(item_revenue_in_usd, na.rm = TRUE),
    transactions = n_distinct(event_timestamp),
    avg_price = mean(price_in_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(revenue))

# Category rollup
category_summary <- product_performance |>
  group_by(item_category) |>
  summarise(
    products = n_distinct(item_name),
    revenue = sum(revenue),
    units = sum(units_sold),
    .groups = "drop"
  ) |>
  mutate(
    revenue_pct = revenue / sum(revenue),
    item_category = fct_reorder(item_category, revenue)
  ) |>
  arrange(desc(revenue))

category_summary |>
  gt() |>
  tab_header(title = "Revenue by Product Category") |>
  fmt_number(columns = c(products, units), decimals = 0) |>
  fmt_currency(columns = revenue, currency = "USD") |>
  fmt_percent(columns = revenue_pct, decimals = 1) |>
  cols_label(
    item_category = "Category",
    products = "Products",
    revenue = "Revenue",
    units = "Units Sold",
    revenue_pct = "Revenue Share"
  ) |>
  data_color(
    columns = revenue,
    method = "numeric",
    palette = c("#e3f2fd", "#1565c0")
  ) |>
  opt_stylize(style = 1)
```

## Analysis 7: User Segmentation

```{r}
#| eval: false

# ── New vs returning user behaviour ───────────────────────────────────────────
user_segments <- ga4_params |>
  filter(key == "ga_session_number") |>
  mutate(
    session_number = as.integer(param_value),
    user_type = if_else(session_number == 1, "New User", "Returning User")
  ) |>
  group_by(user_pseudo_id, user_type) |>
  summarise(
    sessions = n_distinct(param_value),
    .groups = "drop"
  ) |>
  # Join to purchase data
  left_join(
    ga4_events |>
      filter(event_name == "purchase") |>
      group_by(user_pseudo_id) |>
      summarise(
        purchases = n(),
        revenue = sum(ecommerce_purchase_revenue_in_usd, na.rm = TRUE),
        .groups = "drop"
      ),
    by = "user_pseudo_id"
  ) |>
  mutate(
    purchased = !is.na(purchases),
    purchases = replace_na(purchases, 0L),
    revenue = replace_na(revenue, 0)
  ) |>
  group_by(user_type) |>
  summarise(
    total_users = n(),
    buyers = sum(purchased),
    cvr = buyers / total_users,
    total_revenue = sum(revenue),
    avg_revenue = mean(revenue),
    total_sessions = sum(sessions),
    avg_sessions = mean(sessions),
    .groups = "drop"
  )

user_segments |>
  gt() |>
  tab_header(
    title = "New vs Returning User Behaviour",
    subtitle = "Session and revenue analysis via duckplyr"
  ) |>
  fmt_number(columns = c(total_users, buyers, total_sessions), decimals = 0) |>
  fmt_percent(columns = cvr, decimals = 1) |>
  fmt_currency(columns = c(total_revenue, avg_revenue), currency = "USD") |>
  fmt_number(columns = avg_sessions, decimals = 1) |>
  cols_label(
    user_type = "User Type",
    total_users = "Users",
    buyers = "Buyers",
    cvr = "CVR",
    total_revenue = "Total Revenue",
    avg_revenue = "Avg. Revenue/User",
    total_sessions = "Sessions",
    avg_sessions = "Avg. Sessions"
  ) |>
  opt_stylize(style = 1)
```

# **Performance Comparison**

```{r}
#| eval: false
#install.packages("bench")
#install.packages("duckplyr")
#install.packages("dplyr")
library(bench)
library(duckplyr)
library(dplyr)

# Load data both ways for fair comparison
ga4_tibble <- readRDS("data/ga4_raw.rds")
ga4_duck <- duckplyr::df_from_parquet("data/ga4_flat.parquet")

# ── Benchmark the same query both ways ───────────────────────────────────────
bench::mark(
  dplyr = {
    ga4_tibble |>
      filter(event_name == "purchase") |>
      group_by(event_date) |>
      summarise(
        revenue = sum(ecommerce_purchase_revenue_in_usd, na.rm = TRUE),
        n = n(),
        .groups = "drop"
      )
  },
  duckplyr = {
    ga4_duck |>
      filter(event_name == "purchase") |>
      group_by(event_date) |>
      summarise(
        revenue = sum(ecommerce_purchase_revenue_in_usd, na.rm = TRUE),
        n = n(),
        .groups = "drop"
      )
  },
  iterations = 5,
  check = FALSE
) |>
  select(expression, median, mem_alloc) |>
  mutate(
    median = as.character(round(as.numeric(median), 3)),
    mem_alloc = as.character(mem_alloc)
  ) |>
  gt::gt() |>
  gt::tab_header(title = "dplyr vs duckplyr Performance") |>
  gt::opt_stylize(style = 1)
```

# **When to Use Each Tool**   
```{r}
#| echo: false
tibble::tribble(
  ~Situation                            , ~Best_Tool           , ~Why                  ,
  "Data fits in RAM, quick exploration" , "dplyr"              , "Familiar, no setup"  ,
  "Data fits in RAM, need more speed"   , "duckplyr"           , "Same code, faster"   ,
  "Data too large for RAM"              , "duckplyr"           , "Streams from disk"   ,
  "Reading Parquet files"               , "duckplyr"           , "df_from_parquet()"   ,
  "Need SQL window functions"           , "duckplyr or DuckDB" , "Both handle it"      ,
  "Complex multi-table SQL joins"       , "DuckDB (SQL)"       , "SQL clearer here"    ,
  "Sharing code with Python users"      , "DuckDB + Arrow"     , "Parquet interop"     ,
  "Teaching beginners"                  , "duckplyr"           , "No new syntax"       ,
  "Production pipeline"                 , "duckplyr"           , "Speed + simplicity"  ,
  "Custom R functions in mutate()"      , "dplyr fallback"     , "duckplyr falls back"
) |>
  gt::gt() |>
  gt::tab_header(
    title = "Tool Selection Guide",
    subtitle = "dplyr vs duckplyr vs DuckDB"
  ) |>
  gt::cols_label(
    Situation = "Situation",
    Best_Tool = "Best Tool",
    Why = "Reason"
  ) |>
  gt::tab_style(
    style = gt::cell_fill(color = "#e8f5e9"),
    locations = gt::cells_body(rows = Best_Tool == "duckplyr")
  ) |>
  gt::opt_stylize(style = 1)
```