---
title: "Data Acquisition Guide for GA4 Public Data"
subtitle: "90 days of GA4 data in multiple formats"
description: "This guide walks through the complete process of downloading
  three months of GA4 public event data from BigQuery and saving it in
  multiple formats — raw RDS, flat Parquet, and Hive-partitioned Parquet —
  so that all subsequent analyses run entirely from local files without
  requiring a BigQuery connection."
categories:
  - Data Acquisition
  - GA4
  - Parquet
  - Hive Partitioning
  - Rectangling
  - bigrquery
  - Arrow
  - DBI
order: 2
author: "Jae Jung"
date: today
format: 
    html:
        toc: true
        toc-depth: 3
        toc-expand: 1
        code-fold: show
        toc-location: right-body
        number-sections: true
        #embed-resources: true
execute:
  warning: false
  message: false
  echo: fenced
  cache: true
  freeze: auto
---

## Overview {.unnumbered}

This notebook walks through the **one-time process** of downloading three
months of GA4 public event data from Google BigQuery and saving it in
multiple local formats. Once this process completes, every other notebook
in this course loads data from your local machine — no BigQuery connection,
no query costs, no waiting.

::: {.callout-important}
## Run This Notebook Only Once

The code in this notebook downloads data from BigQuery. Once it completes
successfully, **do not run it again**. All subsequent notebooks load from
your local files using the patterns shown at the end of this guide.

Total download time: approximately **20–40 minutes** depending on your
internet connection.
:::

### What You Will Have When This Is Done

After running this notebook you will have the following folder structure
in your project directory:

```
data/
├── ga4_raw/                         ← Raw nested .rds (one per day, 90 files)
│   ├── ga4_20201101.rds
│   ├── ga4_20201102.rds
│   └── ...
├── ga4_parquet/                     ← Flat Parquet (one per day, 90 files)
│   ├── ga4_20201101.parquet
│   ├── ga4_20201102.parquet
│   └── ...
├── ga4_parquet_params/              ← Event params Parquet (one per day)
│   ├── params_20201101.parquet
│   └── ...
├── ga4_parquet_items/               ← Items Parquet (purchase days only)
│   ├── items_20201101.parquet
│   └── ...
├── ga4_partitioned/                 ← Hive-partitioned events dataset
│   ├── year=2020/
│   │   ├── month=11/
│   │   └── month=12/
│   └── year=2021/
│       └── month=1/
├── ga4_params_partitioned/          ← Hive-partitioned params dataset
└── ga4_items_partitioned/           ← Hive-partitioned items dataset
```

### Why Three Formats?

| Format | Where Used in This Course |
|--------|--------------------------|
| `ga4_raw/*.rds` | Teaching rectangling — preserves nested list-columns |
| `ga4_parquet/*.parquet` | Single-day analyses, learning `read_parquet()` |
| `ga4_partitioned/` | Multi-day analyses, learning `open_dataset()` |
| `ga4_params_partitioned/` | Event parameter analysis with Arrow and DuckDB |
| `ga4_items_partitioned/` | E-commerce and product analysis |

---

## Prerequisites

### Required Packages

```{r}
#| label: install-packages
#| eval: false
#| cache: false

# Run this chunk once if you have not installed these packages yet.
# After installing, you do not need to run it again.
install.packages(c(
  "bigrquery", # BigQuery connection and querying
  "DBI", # Standard database interface
  "dplyr", # Data manipulation
  "tidyr", # Rectangling nested columns
  "purrr", # Iteration over lists and vectors
  "arrow", # Parquet read/write and Arrow datasets
  "lubridate", # Date and time handling
  "glue", # String interpolation
  "fs", # File system operations
  "cli", # Formatted console messages
  "stringr", # String manipulation
  "scales" # Number formatting
))
```

```{r}
#| label: load-packages
#| cache: false

library(bigrquery)
library(DBI)
library(dplyr)
library(tidyr)
library(purrr)
library(arrow)
library(lubridate)
library(glue)
library(fs)
library(cli)
library(stringr)
library(scales)
```

### Google Cloud Platform Account Setup

::: {.callout-important}
## Action Required: GCP Account Setup

You need a Google Cloud Platform account before running any code in this
notebook. Follow these steps if you do not have one:

**Step 1 — Create an account**
Go to [console.cloud.google.com](https://console.cloud.google.com) and
sign in with your Google account. A Gmail account works.

**Step 2 — Create a project**

- Click the project selector dropdown at the top of the page
- Click **New Project**
- Give it any name (e.g., `"marketing-analytics"`)
- Click **Create** and wait for the project to be created

**Step 3 — Find your Project ID**

- Your Project ID appears under the project name in the dashboard
- It looks like: `"marketing-analytics-391408"`
- It is **not** the same as the project name — it includes numbers
- Copy it — you will paste it into the code below

**Step 4 — Enable the BigQuery API**

- Go to: [console.cloud.google.com/apis/library/bigquery.googleapis.com](https://console.cloud.google.com/apis/library/bigquery.googleapis.com)
- Confirm your new project is selected in the top navigation bar
- Click **Enable**
- Wait approximately 60 seconds

**Step 5 — Set up billing**

- Go to: [console.cloud.google.com/billing](https://console.cloud.google.com/billing)
- Link a credit card to your account
- **Cost**: Querying the public GA4 dataset costs approximately
  \$0.00–\$0.05 for the entire 90-day download. Google also provides
  \$300 in free credits to all new accounts.
:::

### Set Your Project ID

::: {.callout-warning}
## You Must Complete This Step Before Running Anything Else

Replace `"your-actual-project-id"` in the code chunk below with your
real GCP Project ID. The notebook will stop immediately with a clear
error message if you forget.
:::

```{r}
#| label: set-project-id
#| eval: false

# ── Replace this string with your actual GCP Project ID ──────────────────────
# Example: "marketing-analytics-391408"
BILLING_PROJECT <- "your-actual-project-id" # ← CHANGE THIS LINE

# ── Validate ──────────────────────────────────────────────────────────────────
if (BILLING_PROJECT == "your-actual-project-id") {
  stop(
    "\n",
    "━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━\n",
    "  ACTION REQUIRED\n",
    "\n",
    "  You must replace 'your-actual-project-id' with your\n",
    "  real GCP Project ID before running this notebook.\n",
    "\n",
    "  Find it at: console.cloud.google.com\n",
    "  It looks like: 'my-project-name-391408'\n",
    "━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━\n"
  )
}

if (nchar(trimws(BILLING_PROJECT)) == 0) {
  stop("BILLING_PROJECT cannot be an empty string.")
}

cli::cli_alert_success("Billing project set to: {.val {BILLING_PROJECT}}")
```

---

## Connect to BigQuery

### Authenticate with Google

```{r}
#| label: authenticate
#| eval: false
#| cache: false

# This opens a browser window asking you to sign in with your Google account.
# After signing in once, your credentials are cached — you will not need
# to do this every time you open R.
#
# use_oob = TRUE works reliably in Positron and RStudio when the
# automatic browser redirect does not complete correctly.

bq_auth(use_oob = TRUE)

# Confirm which account you are authenticated as
cli::cli_alert_info("Authenticated as: {.val {bq_user()}}")
```

::: {.callout-note}
## What Does `bq_auth()` Do?

`bq_auth()` handles the OAuth2 authentication handshake between R and
Google. It opens your browser, asks you to grant R permission to access
your Google Cloud account, and then caches the resulting token locally.
On subsequent R sessions, the cached token is used automatically — you
will only see the browser prompt on your first run.
:::

### Establish the Connection

```{r}
#| label: connect
#| eval: false

# ── Connect to the public GA4 dataset ─────────────────────────────────────────
# project  = where the PUBLIC data lives (always "bigquery-public-data")
# dataset  = the specific GA4 dataset name (must be exact — case sensitive)
# billing  = YOUR project (this is what gets charged for query processing)

con <- dbConnect(
  bigrquery::bigquery(),
  project = "bigquery-public-data",
  dataset = "ga4_obfuscated_sample_ecommerce",
  billing = BILLING_PROJECT
)

# ── Verify the connection ─────────────────────────────────────────────────────
available_tables <- dbListTables(con)

cli::cli_alert_success(
  "Connected. {.val {length(available_tables)}} tables available."
)
cli::cli_alert_info(
  "Date range: {.val {head(available_tables, 1)}} to
  {.val {tail(available_tables, 1)}}"
)
```

::: {.callout-note}
## About the Public Dataset

The dataset `ga4_obfuscated_sample_ecommerce` is provided by Google and
contains real event data from the
[Google Merchandise Store](https://shop.merch.google) — an actual
e-commerce site selling Google-branded products. User identifiers have
been obfuscated for privacy. The data covers **November 2020 through
January 2022**, stored as one BigQuery table per day in the format
`events_YYYYMMDD`.
:::

---

## Understand the Data

Before downloading 90 days of data, take a moment to understand the
structure and volume of what you are working with.

### Schema Inspection

```{r}
#| label: inspect-schema
#| eval: false

# ── Inspect the schema of a single day ────────────────────────────────────────
# tbl() creates a lazy reference — nothing is downloaded yet
sample_tbl <- tbl(con, "events_20201101")

# glimpse() fetches a small sample to show column names and types
glimpse(sample_tbl)
```

::: {.callout-note}
## What You Will See in the Schema

The schema shows several column types that are new if you are used to
flat CSV files:

| Column | Type | Notes |
|--------|------|-------|
| `event_date` | `STRING` | Format `"YYYYMMDD"` — not a date type |
| `event_timestamp` | `INT64` | Microseconds since Unix epoch |
| `event_params` | `RECORD REPEATED` | Array of key-value structs |
| `device` | `RECORD` | Nested struct with device attributes |
| `geo` | `RECORD` | Nested struct with location attributes |
| `items` | `RECORD REPEATED` | Array of purchased item structs |

These nested columns are why we need the rectangling techniques taught
in the previous notebook.
:::

### Check Row Counts Per Day

```{r}
#| label: check-row-counts
#| eval: false

# ── Sample a few days to understand data volume ────────────────────────────────
sample_dates <- c(
  "events_20201101", # Start of dataset
  "events_20201201", # One month in
  "events_20210101" # Two months in
)

day_sizes <- sample_dates |>
  map_df(\(tbl_name) {
    n <- tbl(con, tbl_name) |>
      summarise(n = n()) |>
      collect() |>
      pull(n)
    tibble(
      table = tbl_name,
      date = ymd(str_extract(tbl_name, "\\d{8}")),
      row_count = n
    )
  })

day_sizes |>
  mutate(row_count = scales::comma(row_count)) |>
  knitr::kable(
    col.names = c("Table", "Date", "Row Count"),
    caption = "Sample row counts by day"
  )

cli::cli_alert_info(
  "Estimated total rows for 90 days: ~{scales::comma(mean(day_sizes$row_count) * 90)}"
)
```

### Plan the Download

```{r}
#| label: plan-download
#| eval: false

# ── Define the three months to download ───────────────────────────────────────
# November 2020, December 2020, January 2021
download_dates <- seq.Date(
  from = as.Date("2020-11-01"),
  to = as.Date("2021-01-31"),
  by = "day"
)

# Format as BigQuery table name suffixes (YYYYMMDD)
date_strings <- format(download_dates, "%Y%m%d")
table_names <- paste0("events_", date_strings)

# ── Cross-check against tables that actually exist in BigQuery ────────────────
target_tables <- table_names[table_names %in% available_tables]
missing_tables <- table_names[!table_names %in% available_tables]

cli::cli_h2("Download Plan")
cli::cli_alert_info("Dates requested:   {length(date_strings)}")
cli::cli_alert_success("Tables confirmed:  {length(target_tables)}")

if (length(missing_tables) > 0) {
  cli::cli_alert_warning(
    "Tables not found:  {length(missing_tables)} (will be skipped)"
  )
  cat("Missing:", paste(missing_tables, collapse = ", "), "\n")
}

cat("\n")
cli::cli_bullets(c(
  "*" = "Tables to download: {length(target_tables)}",
  "*" = "Estimated rows:     ~{scales::comma(length(target_tables) * 35000)}",
  "*" = "Estimated time:     20-40 minutes",
  "*" = "Estimated storage:  300-500 MB total (all formats)"
))
```

---

## Create the Folder Structure

```{r}
#| label: create-folders

# ── Create all required directories ───────────────────────────────────────────
# fs::dir_create() is safer than base R dir.create() — it does not error
# if the folder already exists, and creates nested folders automatically

dirs_to_create <- c(
  "data",
  "data/ga4_raw",
  "data/ga4_parquet",
  "data/ga4_parquet_params",
  "data/ga4_parquet_items",
  "data/ga4_partitioned",
  "data/ga4_params_partitioned",
  "data/ga4_items_partitioned"
)

walk(dirs_to_create, \(d) {
  fs::dir_create(d, recurse = TRUE)
})

cli::cli_alert_success("All data directories created.")

# ── Confirm they exist ────────────────────────────────────────────────────────
dirs_to_create |>
  keep(fs::dir_exists) |>
  walk(\(d) cli::cli_alert_info("  {d}"))
```

---

## Define Rectangling Helper Functions

GA4 data arrives from BigQuery with nested columns — arrays of structs and
structs of structs. The functions below flatten each type of nested column
into a form that can be saved as Parquet. These are applied to each day's
data immediately after download.

::: {.callout-note}
## Why Define These Before Downloading?

Parquet files cannot store R list-columns (nested data). We must flatten
the nested columns before saving to Parquet. Defining the helper
functions here keeps the download loop clean and readable.
If you need a refresher on why GA4 data is nested and how rectangling
works, refer to the previous notebook in this series.
:::

::: {.callout-note}
## Rectangling: A Brief Appearance

The helper functions below flatten GA4's nested columns before
saving to Parquet. These techniques — `unnest_wider()`,
`unnest_longer()`, and `coalesce()` — are covered in depth in a
dedicated site:

**[Rectangling Hierarchical Data with GA4](https://jaejungca.github.io/ch23-hierarchical_data/hierarchical-data-0.html)**

If the code below looks unfamiliar, visit that site first and
return here when you are ready.
:::

### Struct Flattening Helper

```{r}
#| label: helper-flatten-structs

#' Flatten all struct (RECORD) columns in a GA4 day's data
#'
#' Applies unnest_wider() to device, geo, traffic_source, and ecommerce.
#' Also adds derived date/time columns useful for partitioning and analysis.
#'
#' @param data A tibble downloaded from one GA4 BigQuery table
#' @return A flat tibble with no list-columns from struct fields

flatten_structs <- function(data) {
  # Identify which struct columns are present
  # (not all days have all columns)
  struct_cols <- c("device", "geo", "traffic_source", "ecommerce")
  present <- struct_cols[struct_cols %in% names(data)]

  result <- data |>
    # Remove list-columns that cannot go to Parquet
    # (event_params and items are handled by separate functions)
    select(-any_of(c("event_params", "items")))

  # Unnest each present struct column
  for (col in present) {
    result <- result |>
      unnest_wider(
        !!rlang::sym(col),
        names_sep = "_" # Prefix: device_category, geo_country, etc.
      )
  }

  # Add derived columns for partitioning and convenience
  result |>
    mutate(
      event_date_parsed = ymd(event_date),
      year = year(event_date_parsed),
      month = month(event_date_parsed),
      day = day(event_date_parsed),
      event_dt = as.POSIXct(
        as.numeric(event_timestamp) / 1e6,
        origin = "1970-01-01",
        tz = "UTC"
      )
    ) |>
    select(-event_date_parsed)
}
```

### Event Params Flattening Helper

```{r}
#| label: helper-flatten-params

#' Flatten the event_params repeated RECORD column
#'
#' GA4 stores event parameters as an array of key-value structs where
#' the value is itself a struct with four typed sub-fields (string, int,
#' float, double). This function applies the three-stage pipeline:
#'   Stage 1: unnest_longer() — one row per parameter
#'   Stage 2: unnest_wider()  — expand key and value sub-fields
#'   Stage 3: coalesce()      — consolidate four value types into one
#'
#' The result is kept in LONG format (one row per parameter per event)
#' because pivot_wider() to fully wide format can produce thousands of
#' columns across 90 days of data.
#'
#' @param data A tibble downloaded from one GA4 BigQuery table
#' @return A long tibble: event_date, event_timestamp, event_name,
#'         user_pseudo_id, key, param_value — or NULL if no params exist

flatten_event_params <- function(data) {
  # Guard: check column exists
  if (!"event_params" %in% names(data)) {
    return(NULL)
  }

  # Guard: check at least some rows have params
  has_params <- !all(map_lgl(data$event_params, is.null))
  if (!has_params) {
    return(NULL)
  }

  result <- data |>
    select(
      event_date,
      event_timestamp,
      event_name,
      user_pseudo_id,
      event_params
    ) |>
    # Stage 1: Expand array — one row per parameter per event
    unnest_longer(event_params) |>
    # Stage 2a: Expand struct — get key and value columns
    unnest_wider(event_params) |>
    # Stage 2b: Expand value sub-struct — get typed value columns
    unnest_wider(value, names_sep = "_") |>
    # Stage 3: Consolidate four typed value columns into one
    mutate(
      param_value = coalesce(
        value_string_value,
        as.character(value_int_value),
        as.character(value_float_value),
        as.character(value_double_value)
      )
    ) |>
    select(
      event_date,
      event_timestamp,
      event_name,
      user_pseudo_id,
      key,
      param_value
    )

  if (nrow(result) == 0) {
    return(NULL)
  }
  result
}
```

### Items Flattening Helper

```{r}
#| label: helper-flatten-items

#' Flatten the items repeated RECORD column
#'
#' The items column contains e-commerce line items as an array of named
#' structs. Unlike event_params, each item struct has named fields
#' (item_id, item_name, price, quantity, etc.) so a simple
#' unnest_longer() + unnest_wider() pipeline is sufficient.
#'
#' @param data A tibble downloaded from one GA4 BigQuery table
#' @return A flat tibble with one row per item per purchase event,
#'         or NULL if no purchase events exist in this day's data

flatten_items <- function(data) {
  # Guard: check column exists
  if (!"items" %in% names(data)) {
    return(NULL)
  }

  # Only purchase events have items
  purchase_data <- data |>
    filter(event_name == "purchase") |>
    select(event_date, event_timestamp, user_pseudo_id, items)

  if (nrow(purchase_data) == 0) {
    return(NULL)
  }

  # Guard: check at least some purchases have items populated
  has_items <- purchase_data |>
    filter(!map_lgl(items, \(x) is.null(x) || nrow(x) == 0))

  if (nrow(has_items) == 0) {
    return(NULL)
  }

  has_items |>
    unnest_longer(items) |> # One row per item per purchase
    unnest_wider(items) # Named item fields become columns
}
```

```{r}
#| label: confirm-helpers

cli::cli_alert_success("All three helper functions are defined and ready.")
cli::cli_bullets(c(
  "v" = "flatten_structs()       — flattens device, geo, traffic_source, ecommerce",
  "v" = "flatten_event_params()  — applies 3-stage pipeline to event_params",
  "v" = "flatten_items()         — expands items array for purchase events"
))
```

---

## Download: Day-by-Day Loop

This is the main download process. Each iteration:

1. Checks if the files already exist (skips if so — safe to re-run)
2. Downloads one day's data from BigQuery
3. Saves the raw nested data as `.rds`
4. Flattens struct columns and saves as Parquet
5. Flattens `event_params` and saves as Parquet
6. Flattens `items` and saves as Parquet
7. Logs the result for the summary report

::: {.callout-warning}
## Do Not Interrupt This Process

Each day's files are saved immediately after downloading. If you must
stop, simply close R and re-run the chunk later — already-downloaded
files are detected and skipped automatically.
:::

```{r}
#| label: download-loop
#| eval: false
#| cache: false

cli::cli_h1("Starting GA4 Download")
cli::cli_alert_info(
  "Downloading {length(target_tables)} tables. Please wait..."
)
cat("\n")

# ── Initialise the download log ────────────────────────────────────────────────
download_log <- vector("list", length(target_tables))
names(download_log) <- target_tables

# ── Main loop ─────────────────────────────────────────────────────────────────
for (i in seq_along(target_tables)) {
  tbl_name <- target_tables[i]
  date_str <- str_extract(tbl_name, "\\d{8}")

  # File paths for each format
  rds_path <- glue("data/ga4_raw/ga4_{date_str}.rds")
  parq_path <- glue("data/ga4_parquet/ga4_{date_str}.parquet")
  params_path <- glue("data/ga4_parquet_params/params_{date_str}.parquet")
  items_path <- glue("data/ga4_parquet_items/items_{date_str}.parquet")

  all_exist <- all(
    fs::file_exists(rds_path),
    fs::file_exists(parq_path),
    fs::file_exists(params_path)
  )

  # ── Skip if already downloaded ─────────────────────────────────────────────
  if (all_exist) {
    cli::cli_alert_info(
      "[{i}/{length(target_tables)}] Skipping {date_str} — already exists"
    )
    download_log[[tbl_name]] <- list(status = "skipped", rows = NA_integer_)
    next
  }

  # ── Download with error handling ───────────────────────────────────────────
  cli::cli_progress_step(
    "[{i}/{length(target_tables)}] Downloading {date_str}..."
  )

  day_result <- tryCatch(
    {
      # ── Download raw data ────────────────────────────────────────────────────
      raw_day <- tbl(con, tbl_name) |> collect()
      n_rows <- nrow(raw_day)

      # ── Save raw nested data as .rds ─────────────────────────────────────────
      # compress = FALSE: fastest read/write — use "gz" to save disk space
      saveRDS(raw_day, rds_path, compress = FALSE)

      # ── Flatten structs and save as Parquet ──────────────────────────────────
      flat_day <- flatten_structs(raw_day)
      write_parquet(flat_day, parq_path, compression = "snappy")

      # ── Flatten event_params and save as Parquet ─────────────────────────────
      params_flat <- flatten_event_params(raw_day)
      if (!is.null(params_flat) && nrow(params_flat) > 0) {
        write_parquet(params_flat, params_path, compression = "snappy")
      }

      # ── Flatten items and save as Parquet ────────────────────────────────────
      items_flat <- flatten_items(raw_day)
      if (!is.null(items_flat) && nrow(items_flat) > 0) {
        write_parquet(items_flat, items_path, compression = "snappy")
      }

      # ── Clean up to free RAM before next iteration ───────────────────────────
      rm(raw_day, flat_day, params_flat, items_flat)
      gc(verbose = FALSE)

      cli::cli_alert_success(
        "  Done: {scales::comma(n_rows)} rows saved in 3 formats"
      )

      list(status = "success", rows = n_rows)
    },
    error = function(e) {
      cli::cli_alert_danger(
        "  FAILED: {date_str} — {conditionMessage(e)}"
      )
      list(status = "error", error = conditionMessage(e), rows = NA_integer_)
    }
  )

  download_log[[tbl_name]] <- day_result

  # ── Brief pause every 10 downloads to avoid rate limiting ─────────────────
  if (i %% 10 == 0 && i < length(target_tables)) {
    cli::cli_alert_info("  Pausing briefly to avoid rate limits...")
    Sys.sleep(3)
  }
}

cli::cli_alert_success("Download loop complete!")
```

---

## Build Partitioned Datasets

Now combine all daily Parquet files into Hive-partitioned datasets. This
is the format used for all multi-day analyses in subsequent notebooks.

::: {.callout-note}
## What Is Hive Partitioning?

Hive partitioning stores data in a folder hierarchy where folder names
encode column values: `year=2020/month=11/part-0.parquet`. When you
query `filter(year == 2020, month == 11)`, Arrow and DuckDB skip every
other folder entirely — they never open those files. This makes
date-filtered queries dramatically faster on large datasets.
:::

### Events Partitioned Dataset

```{r}
#| label: partition-events
#| eval: false
#| cache: false

cli::cli_h2("Building partitioned events dataset")

# ── Find all daily Parquet files ──────────────────────────────────────────────
event_files <- fs::dir_ls("data/ga4_parquet", regexp = "\\.parquet$")

if (length(event_files) == 0) {
  cli::cli_abort(
    "No Parquet files found in data/ga4_parquet/.
    Run the download loop first."
  )
}

cli::cli_alert_info(
  "Found {length(event_files)} daily Parquet files to combine"
)

# ── Open all files as one Arrow dataset and write partitioned ─────────────────
open_dataset(event_files) |>
  write_dataset(
    path = "data/ga4_partitioned",
    format = "parquet",
    partitioning = c("year", "month"),
    compression = "snappy",
    max_rows_per_file = 500000
  )

cli::cli_alert_success(
  "Events partitioned dataset written to data/ga4_partitioned/"
)
```

### Event Params Partitioned Dataset

```{r}
#| label: partition-params
#| eval: false
#| cache: false

cli::cli_h2("Building partitioned event_params dataset")

params_files <- fs::dir_ls(
  "data/ga4_parquet_params",
  regexp = "\\.parquet$"
)

if (length(params_files) == 0) {
  cli::cli_alert_warning(
    "No params Parquet files found. Skipping params partitioning."
  )
} else {
  cli::cli_alert_info(
    "Found {length(params_files)} params files to combine"
  )

  open_dataset(params_files) |>
    mutate(
      year = as.integer(str_sub(event_date, 1L, 4L)),
      month = as.integer(str_sub(event_date, 5L, 6L))
    ) |>
    write_dataset(
      path = "data/ga4_params_partitioned",
      format = "parquet",
      partitioning = c("year", "month"),
      compression = "snappy"
    )

  cli::cli_alert_success(
    "Params partitioned dataset written to data/ga4_params_partitioned/"
  )
}
```

### Items Partitioned Dataset

```{r}
#| label: partition-items
#| eval: false
#| cache: false

cli::cli_h2("Building partitioned items dataset")

items_files <- fs::dir_ls(
  "data/ga4_parquet_items",
  regexp = "\\.parquet$"
)

if (length(items_files) == 0) {
  cli::cli_alert_warning(
    "No items Parquet files found.
    This is expected if no purchase events occurred.
    Skipping items partitioning."
  )
} else {
  cli::cli_alert_info(
    "Found {length(items_files)} items files to combine"
  )

  open_dataset(items_files) |>
    mutate(
      year = as.integer(str_sub(event_date, 1L, 4L)),
      month = as.integer(str_sub(event_date, 5L, 6L))
    ) |>
    write_dataset(
      path = "data/ga4_items_partitioned",
      format = "parquet",
      partitioning = c("year", "month"),
      compression = "snappy"
    )

  cli::cli_alert_success(
    "Items partitioned dataset written to data/ga4_items_partitioned/"
  )
}
```

---

## Verify the Download

Run these checks after the download completes to confirm everything is
correct before closing the connection.

### File Count Check

```{r}
#| label: verify-file-counts

cli::cli_h2("Verification: File Counts")

# Define what we expect in each folder
expected_counts <- tibble::tribble(
  ~folder                   , ~description                         ,
  "data/ga4_raw"            , "Raw nested .rds files"              ,
  "data/ga4_parquet"        , "Flat events Parquet"                ,
  "data/ga4_parquet_params" , "Event params Parquet"               ,
  "data/ga4_parquet_items"  , "Items Parquet (purchase days only)"
)

# Count files actually present
file_counts <- expected_counts |>
  mutate(
    exists = map_lgl(folder, fs::dir_exists),
    n_files = map_int(folder, \(d) {
      if (!fs::dir_exists(d)) {
        return(0L)
      }
      length(fs::dir_ls(d, regexp = "\\.(rds|parquet)$"))
    }),
    status = case_when(
      !exists ~ "folder missing",
      n_files == 0 ~ "empty",
      n_files < 85 ~ glue("⚠ only {n_files} files"),
      n_files >= 85 ~ glue("✓ {n_files} files"),
      TRUE ~ as.character(n_files)
    )
  ) |>
  select(description, folder, n_files, status)

knitr::kable(
  file_counts,
  col.names = c("Dataset", "Folder", "Files", "Status"),
  caption = "File count by folder"
)
```

### Storage Summary

```{r}
#| label: verify-storage

cli::cli_h2("Verification: Storage Used")

all_data_folders <- c(
  "data/ga4_raw",
  "data/ga4_parquet",
  "data/ga4_parquet_params",
  "data/ga4_parquet_items",
  "data/ga4_partitioned",
  "data/ga4_params_partitioned",
  "data/ga4_items_partitioned"
)

storage_summary <- all_data_folders |>
  keep(fs::dir_exists) |>
  map_df(\(d) {
    files <- fs::dir_ls(d, recurse = TRUE, regexp = "\\.(rds|parquet)$")
    size_mb <- sum(as.numeric(fs::file_size(files))) / 1024^2
    tibble(
      folder = d,
      n_files = length(files),
      size_mb = round(size_mb, 1)
    )
  })

storage_summary |>
  mutate(size_mb = glue("{size_mb} MB")) |>
  knitr::kable(
    col.names = c("Folder", "Files", "Size"),
    caption = "Storage used by folder"
  )

total_mb <- sum(storage_summary$size_mb)
cli::cli_alert_info(
  "Total storage used: {round(total_mb / 1024, 2)} GB
  ({scales::comma(total_mb)} MB)"
)
```

### Partition Structure Check

```{r}
#| label: verify-partitions
#| eval: false

cli::cli_h2("Verification: Partition Structure")

if (fs::dir_exists("data/ga4_partitioned")) {
  ga4_check <- open_dataset("data/ga4_partitioned")

  cat("Schema:\n")
  print(ga4_check$schema)
  cat("\n")

  partition_summary <- ga4_check |>
    count(year, month, name = "rows") |>
    collect() |>
    arrange(year, month) |>
    mutate(
      month_name = month.abb[month],
      period = glue("{month_name} {year}"),
      rows = scales::comma(rows)
    ) |>
    select(period, year, month, rows)

  knitr::kable(
    partition_summary,
    col.names = c("Period", "Year", "Month", "Rows"),
    caption = "Row counts by partition"
  )

  total_rows <- ga4_check |>
    summarise(n = n()) |>
    collect() |>
    pull(n)

  cli::cli_alert_success(
    "Total rows across all partitions: {scales::comma(total_rows)}"
  )
} else {
  cli::cli_alert_warning(
    "Partitioned dataset not found. Run the partitioning chunks above."
  )
}
```

### Sample Data Check

```{r}
#| label: verify-sample-data
#| eval: false

cli::cli_h2("Verification: Sample Data")

# ── Core events ───────────────────────────────────────────────────────────────
cat("── Core events (first 5 rows, November 2020) ──────────────────────\n")
if (fs::dir_exists("data/ga4_partitioned")) {
  open_dataset("data/ga4_partitioned") |>
    filter(year == 2020, month == 11) |>
    select(
      event_date,
      event_name,
      user_pseudo_id,
      device_category,
      geo_country
    ) |>
    head(5) |>
    collect() |>
    knitr::kable()
}

# ── Event params ──────────────────────────────────────────────────────────────
cat("\n── Event params (first 5 rows) ─────────────────────────────────────\n")
if (fs::dir_exists("data/ga4_params_partitioned")) {
  open_dataset("data/ga4_params_partitioned") |>
    filter(year == 2020, month == 11) |>
    head(5) |>
    collect() |>
    knitr::kable()
}

# ── Items ─────────────────────────────────────────────────────────────────────
cat("\n── Items (first 5 rows) ─────────────────────────────────────────────\n")
if (fs::dir_exists("data/ga4_items_partitioned")) {
  open_dataset("data/ga4_items_partitioned") |>
    head(5) |>
    collect() |>
    knitr::kable()
}
```

---

## Download Summary Report

```{r}
#| label: download-summary
#| eval: false

cli::cli_h1("Download Summary")

if (exists("download_log") && length(download_log) > 0) {
  # ── Build log data frame ──────────────────────────────────────────────────
  log_df <- download_log |>
    imap_df(\(x, nm) {
      tibble(
        table = nm,
        status = x$status,
        rows = x$rows %||% NA_integer_
      )
    }) |>
    mutate(
      date = ymd(str_extract(table, "\\d{8}")),
      month_label = glue("{month.abb[month(date)]} {year(date)}")
    )

  # ── Status summary ────────────────────────────────────────────────────────
  cat("Status Summary:\n")
  log_df |>
    count(status, name = "tables") |>
    knitr::kable(col.names = c("Status", "Tables"))

  # ── Monthly summary ───────────────────────────────────────────────────────
  cat("\nRows by Month:\n")
  log_df |>
    filter(status == "success") |>
    group_by(month_label) |>
    summarise(
      days_downloaded = n(),
      total_rows = sum(rows, na.rm = TRUE),
      avg_rows_per_day = round(mean(rows, na.rm = TRUE)),
      .groups = "drop"
    ) |>
    mutate(
      total_rows = scales::comma(total_rows),
      avg_rows_per_day = scales::comma(avg_rows_per_day)
    ) |>
    knitr::kable(
      col.names = c("Month", "Days", "Total Rows", "Avg Rows/Day")
    )

  # ── Flag any failures ─────────────────────────────────────────────────────
  failed <- log_df |> filter(status == "error")

  if (nrow(failed) > 0) {
    cli::cli_alert_danger("{nrow(failed)} tables failed:")
    failed |>
      select(table, status) |>
      print()
    cli::cli_alert_info(
      "Re-run the download loop — it skips successful days automatically."
    )
  } else {
    cli::cli_alert_success("All tables downloaded successfully!")
  }
}
```

---

## Disconnect from BigQuery

```{r}
#| label: disconnect
#| eval: false
#| cache: false

# Close the BigQuery connection — no longer needed after download
dbDisconnect(con)
cli::cli_alert_success("BigQuery connection closed.")
cli::cli_alert_info(
  "You will not need to reconnect to BigQuery for the rest of this course."
)
```

---

## How to Load Data in Future Sessions

::: {.callout-tip}
## Quick Reference — Copy This Into Every Future Notebook

Paste this code block at the top of each subsequent notebook to load
all three datasets:

```r
library(arrow)
library(dplyr)

ga4_events <- open_dataset("data/ga4_partitioned")
ga4_params <- open_dataset("data/ga4_params_partitioned")
ga4_items  <- open_dataset("data/ga4_items_partitioned")
```

That is all you need. No BigQuery, no authentication, no waiting.
:::

### The Five Loading Patterns

```{r}
#| label: loading-patterns
#| eval: false

library(arrow)
library(dplyr)
library(duckdb)
library(duckplyr)

# ── Pattern 1: Partitioned dataset — recommended for most analyses ─────────────
# Lazy: nothing loaded into RAM until collect()
# Fast: Arrow skips irrelevant year/month folders automatically
ga4_events <- open_dataset("data/ga4_partitioned")
ga4_params <- open_dataset("data/ga4_params_partitioned")
ga4_items <- open_dataset("data/ga4_items_partitioned")

# Query a specific month
november_events <- ga4_events |>
  filter(year == 2020, month == 11) |>
  select(event_date, event_name, user_pseudo_id, device_category) |>
  collect()

# ── Pattern 2: Single day flat Parquet ────────────────────────────────────────
# Eager: loads immediately into RAM as a tibble
# Use for: single-day analysis, quick exploration
day_data <- read_parquet("data/ga4_parquet/ga4_20201101.parquet")

# ── Pattern 3: Raw nested .rds ────────────────────────────────────────────────
# Use for: teaching and practising rectangling techniques
raw_day <- readRDS("data/ga4_raw/ga4_20201101.rds")
glimpse(raw_day) # List-columns are still nested — ready for rectangling

# ── Pattern 4: DuckDB on partitioned Parquet ──────────────────────────────────
# Use for: SQL queries, window functions, complex multi-table joins
con <- dbConnect(duckdb::duckdb())
duckdb_register_arrow(con, "events", open_dataset("data/ga4_partitioned"))
duckdb_register_arrow(
  con,
  "params",
  open_dataset("data/ga4_params_partitioned")
)
duckdb_register_arrow(con, "items", open_dataset("data/ga4_items_partitioned"))

DBI::dbGetQuery(
  con,
  "
  SELECT event_name, COUNT(*) AS n
  FROM events
  GROUP BY event_name
  ORDER BY n DESC
  LIMIT 10
"
)

DBI::dbDisconnect(con, shutdown = TRUE)

# ── Pattern 5: duckplyr on Parquet ────────────────────────────────────────────
# Use for: dplyr syntax with DuckDB speed on large data
ga4_duck <- duckplyr::df_from_parquet(
  fs::dir_ls("data/ga4_parquet", regexp = "\\.parquet$")
)

ga4_duck |>
  count(event_name, sort = TRUE) |>
  head(10)
```

---

## Troubleshooting

```{r}
#| label: troubleshooting
#| eval: false

# ── Problem 1: "notFound" error ───────────────────────────────────────────────
# Symptom: Error in bq_post(): Request couldn't be served. [notFound]
# Causes and fixes:

# Cause A: BILLING_PROJECT is wrong
cat("Current billing project:", BILLING_PROJECT, "\n")
bq_projects() # Your project MUST appear in this list

# Cause B: BigQuery API not enabled on your project
# Fix: Go to console.cloud.google.com/apis/library/bigquery.googleapis.com
#      select your project and click Enable

# Cause C: Dataset name has a typo
# The correct name is exactly:
# "ga4_obfuscated_sample_ecommerce"
# Check available datasets:
bq_project_datasets("bigquery-public-data") |>
  as_tibble() |>
  dplyr::filter(stringr::str_detect(dataset, "ga4"))

# ── Problem 2: Integer overflow warning ───────────────────────────────────────
# Symptom: NAs produced by integer overflow
# Cause: event_timestamp is INT64, too large for R's 32-bit integer
# Fix: collect() via tbl() handles this automatically through Arrow
# If using bq_table_download() directly, add: bigint = "numeric"

# ── Problem 3: Download stalls or times out ───────────────────────────────────
# Fix: The loop saves each day before moving to the next
# Simply re-run the download chunk — already-saved files are skipped
already_saved <- fs::dir_ls("data/ga4_parquet") |>
  fs::path_file() |>
  str_extract("\\d{8}") |>
  na.omit()

still_needed <- date_strings[!date_strings %in% already_saved]
cli::cli_alert_info(
  "{length(still_needed)} tables still need downloading"
)

# ── Problem 4: "Cannot allocate vector" — out of memory ──────────────────────
# Cause: One day's data is larger than available RAM
# Fix: Download fewer columns per day
tbl(con, "events_20201101") |>
  select(
    event_date,
    event_timestamp,
    event_name,
    user_pseudo_id,
    device,
    geo,
    traffic_source
  ) |>
  collect() # Excludes event_params and items — much smaller

# ── Problem 5: Parquet write error with list-columns ─────────────────────────
# Symptom: Error writing Parquet — type not supported
# Cause: Trying to write a tibble that still has list-columns
# Fix: The flatten_structs() function removes event_params and items
# before writing. Check that it ran without error.

# ── Problem 6: Partitioned dataset reads wrong columns ───────────────────────
# Symptom: year and month columns show as integers but filtering fails
# Fix: Confirm the partition columns exist in the schema
open_dataset("data/ga4_partitioned")$schema
# year and month should appear as int32 or int64
```

---

## Session Info

```{r}
#| label: session-info
#| code-fold: true
#| code-summary: "Show session info"
#| cache: false
#| eval: false

sessioninfo::session_info()
```

